### 2025-04-03 _archive/ellipsis-projection.qmd
Initial posterior work done, missing causal section

---
title: Ellipsis Projection
author: Daniel Posmik
date: today
format: pdf
toc: false
number_sections: true
---

```{r setup, include=FALSE}
# Set up knit environment
knitr::opts_chunk$set(echo = F)
knitr::opts_chunk$set(error = F)
knitr::opts_chunk$set(warning = F)
knitr::opts_chunk$set(message = F)
```

## Simulation of Points on an Elliptic Curve

First, we set up our parameters and generate points on an elliptic curve.

```{r}
# Set random seed for reproducibility
set.seed(123)

# Parameters for the elliptic curve
a <- 3  # semi-major axis
b <- 1  # semi-minor axis

# Generate 5 points on an elliptic curve
n <- 20
theta <- seq(0, 2*pi, length.out = n+1)[1:n]  # equally spaced angles

# Generate points on the perfect ellipse
x_ellipse <- a * cos(theta)
y_ellipse <- b * sin(theta)

# Add Gaussian noise N(0,1) to create observed points
noise_sd <- 0.2
x_observed <- x_ellipse + rnorm(n, 0, noise_sd)
y_observed <- y_ellipse + rnorm(n, 0, noise_sd)

# Generate uniform(0,1) values for the z coordinate
z_values <- runif(n)

# Create the data frame
data <- data.frame(
  x = x_observed,
  y = y_observed,
  z = z_values
)
```

## Projection onto the Ellipse

Here we project the noisy points back onto the ellipse, considering all three dimensions.

```{r}
# Project a point onto the ellipse and return coordinates and angles
project_to_ellipse <- function(x, y, z, a, b) {
  # Starting guess for theta (x-y plane angle)
  theta_guess <- atan2(y/b, x/a)

  # Function to minimize: squared distance from (x,y,z) to a point on the ellipse
  distance_function <- function(theta) {
    ellipse_x <- a * cos(theta)
    ellipse_y <- b * sin(theta)
    ellipse_z <- 0 # Since our ellipse only lives in x,y space
    return(sqrt((x - ellipse_x)^2 + (y - ellipse_y)^2 + (z - ellipse_z)^2))
  }

  # Optimize to find the best theta
  result <- optimize(distance_function, c(0, 2*pi))
  theta_optimal <- result$minimum

  # Calculate projected coordinates
  x_proj <- a * cos(theta_optimal)
  y_proj <- b * sin(theta_optimal)
  z_proj <- 0

  # Calculate distance between original and projected point
  dist <- sqrt((x - x_proj)^2 + (y - y_proj)^2 + (z - z_proj)^2)

  # Calculate the second angle (phi) - elevation from x-y plane
  # This is the angle between the x-y plane and the line to the original point
  phi <- atan2(z, sqrt(x^2 + y^2))

  # Return the projected point coordinates, angles, and distance
  return(list(
    x_proj = x_proj,
    y_proj = y_proj,
    z_proj = z_proj,
    theta = theta_optimal,  # Azimuthal angle in x-y plane
    phi = phi,              # Elevation angle from x-y plane
    distance = dist
  ))
}

# Apply the projection to each point
projection_results <- lapply(1:n, function(i) {
  project_to_ellipse(data$x[i], data$y[i], data$z[i], a, b)
})

# Extract results into data frame
data$x_projected <- sapply(projection_results, function(res) res$x_proj)
data$y_projected <- sapply(projection_results, function(res) res$y_proj)
data$z_projected <- sapply(projection_results, function(res) res$z_proj)
data$projection_theta <- sapply(projection_results, function(res) res$theta)
data$projection_phi <- sapply(projection_results, function(res) res$phi)
data$projection_distance <- sapply(projection_results, function(res) res$distance)

# Print the data frame with projection information
head(data)
```

## 2D Visualization

```{r, fig.width=8, fig.height=8}
library(ggplot2)

# Create a dense set of points for drawing the perfect ellipse
theta_dense <- seq(0, 2*pi, length.out = 100)
ellipse_points <- data.frame(
  x = a * cos(theta_dense),
  y = b * sin(theta_dense)
)

# Create the plot
p <- ggplot() +
  # Draw the perfect ellipse
  geom_path(data = ellipse_points, aes(x = x, y = y), color = "gray50", linetype = "dashed") +
  # Draw the observed points with color based on z-value
  geom_point(data = data, aes(x = x, y = y, color = z), size = 3) +
  # Add a color scale for the z-values
  scale_color_viridis_c(name = "Z Value") +
  # Draw the projected points
  geom_point(data = data, aes(x = x_projected, y = y_projected), color = "purple", shape = 1, size = 3) +
  # Draw lines connecting observed points to their projections
  geom_segment(data = data, 
               aes(x = x, y = y, xend = x_projected, yend = y_projected),
               linetype = "dotted") +
  # Add labels and theme
  labs(title = "Elliptic Curve Simulation and Projection (X-Y Plane View)",
       subtitle = "Original points (colored by z-value) and their projections (purple circles)",
       x = "X", y = "Y") +
  theme_minimal() +
  coord_equal()  # Equal aspect ratio for proper ellipse visualization

# Display the plot
print(p)
```

## Summary of Projection Metrics

```{r}
# Create summary table
summary_table <- data.frame(
  Point = 1:n,
  Original_X = data$x,
  Original_Y = data$y,
  Original_Z = data$z,
  Projected_X = data$x_projected,
  Projected_Y = data$y_projected,
  Projected_Z = data$z_projected,
  Theta_Angle = data$projection_theta,  # Azimuthal angle in x-y plane
  Phi_Angle = data$projection_phi,      # Elevation angle from x-y plane
  Projection_Distance = data$projection_distance
)

# Print the summary table
head(summary_table)
```

# **Code Appendix**

```{r, ref.label = knitr::all_labels()}
#| echo: true
#| eval: false
```

### 2025-04-03 output/paper.tex
Initial posterior work done, missing causal section

\documentclass[twoside,11pt]{article}

\usepackage{paper}
\usepackage{tikz}     
\usepackage{xcolor}
\usepackage{amsmath}
\usetikzlibrary{calc}
\usepackage{graphicx}
\usepackage{subcaption}

\ShortHeadings{
%% short title here
Directional Priors in Manifolds
}{
% last name
Posmik}
\firstpageno{1}

\begin{document}

\title{	Directional Priors in Manifold Learning \\
\vspace{.1in}
PHP2530			
}

\author{ Daniel C. Posmik }

\maketitle
\date{4 }

\section{Introduction} \label{sc:intro}

Manifold learning is dimensionality reduction technique that has proven useful in settings where data are high-dimensional and non-linear. Often, manifold learning algorithms are used when the topological structure of the data are to be preserved in a statistical learning task\footnote{For an introduction, see \citet{Meila2023}}. Within the manifold learning framework, data are assumed to live on a lower dimensional manifold and are corrupted by high-dimensional noise. We say that $D$-dimensional data can be embedded in $d$-dimensional manifold where $d \leq D$ but generally $d \ll D$ .

\begin{figure}
  \begin{center}
    \begin{tikzpicture}[scale=0.85]
      % Define styles
      \tikzset{
        manifold/.style={blue, thick, fill=blue!10},
        noise/.style={gray, thick},
        observed/.style={blue, thick, fill=blue!10, opacity=0.7},
        arrow/.style={->, >=stealth, thick},
        label/.style={font=\normalfont\normalsize}
      }
      % Low-dimensional manifold (left)
      \begin{scope}[xshift=-5cm]
        \draw[manifold] plot[smooth, tension=0.8] coordinates {(-1.5,0) (-0.5,1) (1,1.2) (1.5,0) (1,-1) (-0.5,-1) (-1.5,0)};

        % Draw some points on the manifold
        \node[circle, draw, inner sep=1pt, fill=blue!50] at (-0.7,0.3) {$f$};
        \node[circle, draw, inner sep=1pt, fill=blue!50] at (0.8,0.5) {$f$};
        \node[circle, draw, inner sep=1pt, fill=blue!50] at (0.2,-0.6) {$f$};
      \end{scope}
      % High-dimensional noise (center)
      \begin{scope}
        % Draw noise as radial lines from center point
        \fill (0,0) circle (0.1);
        \foreach \i in {0,10,...,350} {
          \draw[noise] (0,0) -- (\i:1.2+0.3*rnd);
        }
      \end{scope}
      % Observed data (right)
      \begin{scope}[xshift=5cm]
        % First draw the noise pattern
        \fill (0,0) circle (0.1);
        \foreach \i in {0,10,...,350} {
          \draw[noise] (0,0) -- (\i:1.2+0.3*rnd);
        }

        % Then overlay the manifold with transparency
        \draw[observed] plot[smooth, tension=0.8] coordinates {(-1.5,0) (-0.5,1) (1,1.2) (1.5,0) (1,-1) (-0.5,-1) (-1.5,0)};

        % Draw some points on the manifold
        \node[circle, draw, inner sep=1pt, fill=blue!50, opacity=0.7] at (-0.7,0.3) {$f$};
        \node[circle, draw, inner sep=1pt, fill=blue!50, opacity=0.7] at (0.8,0.5) {$f$};
        \node[circle, draw, inner sep=1pt, fill=blue!50, opacity=0.7] at (0.2,-0.6) {$f$};
      \end{scope}
      % Draw the operation symbols
      \node at (-2.5,0) {$+$};
      \draw[arrow] (2,0.5) -- (3,0.5);
      \draw[arrow] (3,-0.5) -- (2,-0.5);
      % Draw the curved arrows for parametrization and embedding
      \draw[arrow, green!50!black, thick] (-4,2.5) to[bend left=30] node[above, font=\normalfont\normalsize] {Parameterization} (4,2.5);
      \draw[arrow, orange, thick] (4,-2.5) to[bend left=30] node[below, font=\normalfont\normalsize] {Embedding} (-4,-2.5);
      % Labels with consistent position and size
      \node[label, text width=3cm, align=center] at (-5,-2) {Low-dimensional manifold};
      \node[label, text width=3cm, align=center] at (0,-2) {High-dimensional noise};
      \node[label, text width=3cm, align=center] at (5,-2) {Observed data};
    \end{tikzpicture}
  \end{center}
  \caption{Manifold Learning: Parameterization vs. Embedding}\label{fig:manifolds}
\end{figure}

Within the principal manifolds framework \citep{Meng2021} -- a replicable and flexible framework for manifold learning -- the process of fitting a manifold to our data contains multiple steps. The key idea is that we fit a $d$-dimensional manifold to our $D$-dimensional by minimizing the sum of squares between our data and the proposed manifold. An important extension to linear dimensionality reduction, i.e. the principal components algorithm (PCA), is that we allow our proposed manifold to preserve underlying topological structure of our data. In a way, manifold learning reduces the dimensionality of data with an explicit focus on the topology of it. We note that -- although certainly intuitive -- this topological structure is not only limited to spatial abstraction, but may be extended to arbitrary dimensions of interest. This framework was pioneered as an extension to the PCA algorithm with curves (\citet{HastieStuetzle1989}, \citet{Tibshirani1992}) and has since found a myriad of applications in higher-dimensional extensions.

We now propose a method for incorporating prior distributional information into the principal manifolds framework.

\section{Background}

Consider a setting where we have fit a manifold $\mathcal{M}_d$ to our $D$-dimensional data by means of minimizing the orthogonal distance between the data and the manifold. We consider this manifold fixed and will not touch on the fitting procedure itself. Given $\mathcal{M}_d$, for each data point, i.e. the row vector $[x_{11} \cdots x_{1D}]^T$, we can now define the point on $\mathcal{M}_d$, say $f\left(\left[x_{11} \cdots x_{1D}\right]^T\right)$. This point minimizes the distance between $x_i$ and $f(x_i)$. We want to stress again that this procedure does not mean we are fitting the manifold to the data, we are simply retrieving the distance-minimizing projection point. We write

$$
\text{arg~min}_{f \in \mathcal{F}} \|x^* - f(x^*)\|_2
$$

where we consider each projection function $f$ to be a member of an arbitrary function space $\mathcal{F}$. We define the distance metric as the $L^2$ distance.

If there exists only one projection point $f(x_i)$ for every $x_i$, every $f \in \mathcal{F}$ is one-to-one and onto ("bijective") mapping. We find it interesting to highlight that the projection functions in the PCA algorithm are inherently bijective, and for inferential purposes, this is a property that is often taken for granted\footnote{This is because a principal axes is a straight line, i.e. neither convex or concave. Although a point's distance to its projection may be co-minimal across $\leq 2$ dimensions, it only has one $f(x_i)$ in one principal axis.}. In a manifold learning framework, this is no longer the case. Albeit highly interesting, due to the limited scope of this paper, we shall treat this scenario as an edge case, reserving rigorous treatment for the blissful times that follow the author's qualifying exam.

Now, given the data $[\{x_i\}_{i=1}^n, \{f(x_i)\}_{i=1}^n]$, we can reparameterize our space into polar coordinates to obtain a vector representation of the collection $f \in \mathcal{F}$. Converting a Cartesian parameterization in space with $D$ dimensions into polar coordinates yields the $d$-dimensional vector $[r_i^*; \theta_{i, 1}, \cdots, \theta_{i, D-1}]$, i.e. one radius $r_i^*$ and a set of $D-1$ angles suffice to characterize each point $x_i$'s location in space.

Recognize that the parameter $r^*$ is not random. This is because it is simply the result from our previous projection distance-minimizing procedure. Usually, polar parameterizations assume that all angles and radii are centered at the origin. Luckily, simple vector addition and subtraction readily generalizes our parameterizations in space. For instance, to obtain the vector from the point $x_i$ and $f(x_i)$, we simply subtract $f(x_i) - x_i$. It is important that the issue of defining the origin is explicitly clarified when dealing with directional information. For simplicity, we will henceforth consider data centered at the origin.

\begin{figure}
  \begin{center}
    \includegraphics[width=0.75\textwidth]{../fig/data-with-noise.png}
  \end{center}
  \caption{Noisy data projected on $\mathcal{M}_3$; the unit sphere in $\mathbb{R}^3$}\label{fig:data-with-noise}
\end{figure}

\section{Directional Priors}

In contrast to the radii, the $\{D-1\}$-dimensional vector of angles, $\boldsymbol{\theta_i} := [\theta_{i, 1}, \cdots, \theta_{i, D-1}]^T$ is random. In simple terms, $\{r_i, \boldsymbol{\theta}_i\}$ is what parameterizes the realized sample $\boldsymbol{X_i} = x_i$ in space. If we draw multiple samples, the random sampling variation in $\mathbf{\theta}_i$ is what captures the randomness. In biomedical applications, such as cancer medicine, we may have reliable prior information (i.e. from previous trials or expert knowledge) on directional trends of malignant growths. Within a Bayesian framework, using our data to update these prior directional information offers a principled, probabilistic solution to complex inference problems in settings where directionality is a key piece of information.

Before we formulate our approach formally, we briefly introduce directional statistics. When we reparameterized our data from Cartesian coordinates into polar coordinates, we did not address the underlying probabilistic transformations. For example, when encoding uncertainty in a spherical setting, it may be naive to parameterize a normal density with support on $\mathbb{R}^1$ since angles are defined in the interval $[0, 2\pi]$. Directional statistics offer well-defined spherical reparameterizations of distributions, such as the normal. Instead of defining a mean vector in $\mathbb{R}^1$, we can define a mean directional vector contained in $[0, 2\pi]$. To account for the infinite support, the wrapped normal is defined on a support of $k\text{mod}(2\pi)$. Henceforth, unless stated otherwise, we will deal with the spherical parameterizations of densities to account for the spherical parameterization of our problem.

\subsection{Motivating Example: 3D Sphere}

We begin our discussion with an example of a 3D sphere. As discussed, this means that we have some simulated noisy data $\{x_i\}_{i=1}^n$ in 3D space. The noisy data is projected on the manifold of choice, i.e. the unit sphere. This yields the data $\{f(x_i)\}_{i=1}^n$. These data are parameterized in a polar coordinate system with $\{r_i, \mathbf{\theta}_i\}$, where the (non-random) radius $\vec{r_i}$ is the vector from the origin to $f(x_i)$. Since we are in 3D space, we have two angles. One angle, $\{\mathbf{\theta}\}_{i=1}^n$, which parameterizes the angle from the origin on the XY-plane. The angle $\{\mathbf{\phi}\}_{i=1}^n$ is the angle between the XY-plane and the point $f(x_i)$.

\begin{figure}
  \begin{center}
    \includegraphics[width=0.75\textwidth]{../fig/projections-from-origin.png}
  \end{center}
  \caption{Polar parameterization with $\{r_i, \theta_i, \phi_i\}$ of projected data}\label{fig:projections-from-origin}
\end{figure}

\subsection{Formulating Directional Priors}

Formulating a directional prior is not trivial. For one, it is important to set a reference point for both the directional prior and the angles. It is important that all angles are measured from the same reference point. For simplicity, we choose the origin. Now, choosing a directional prior in some direction $\mu \in [0, \pi]$ allows\footnote{Note that we choose prior means are only meaningful in $[0, \pi]$ This is because for any prior mean greater than $\pi$, we would consider the closer angle in the opposite direction.} us to incorporate information on where the projected points should be concentrated on the manifold. We only posit a prior on the random set $\{\theta_i\}_{i=1}^n$, expressing some prior distributional belief on the concentrations of points in the XY-plane. In this example, we consider the angles $\{\phi_i\}_{i=1}^n$ to be fixed, making our example univariate. An extension to the multivariate case would be straightforward by using the multivariate analogoues of our likelihood and prior distribution functions.

Our posterior distribution on the random vector $\theta$ can be expressed as follows.

\[
\begin{aligned}
  f_{\theta | \mu} (\theta | \mu) 
  &\propto \left\{ \prod_{i=1}^n f_{\theta_i | \mu}(\theta_i | \mu) \right\} \cdot f_{\mu}(\mu) \\ 
  &\propto \mathcal{L}(\theta | \mu) \cdot f_{\mu}(\mu) 
\end{aligned}
\]

where $\boldsymbol{\mu}$ is our mean vector that we posit a prior on, i.e. $\mu \sim f_{\mu}(\mu)$. To form our likelihood, we rely on the wrapped ("spherical") normal distribution. The density of a wrapped normal is given by

$$
f(\theta|\mu,\sigma^2) = 
\frac{1}{2\pi\sigma^2} \sum_{k=-\infty}^{\infty} \exp\left(-\frac{(\theta-\mu+2\pi k)^2}{2\sigma^2}\right)
$$

with the infinite support reflected in the $\mathit{mod}(k)$ argument. Integrating over $2\pi$ is equivalent to integrating over $\mathbb{R}$ in the canonical parameterization. Now, a disadvantage of using the above form is the infinite series. Deriving moments is not straightforward and requires (inverse) Fourier transforms. Thus, we will use the von Mises distribution, an analogue to the spherical normal that is easier to handle. The density of a von Mises distribution is given by

$$
f(\theta|\mu,\kappa) = \frac{e^{\kappa \cos(\theta-\mu)}}{2\pi I_0(\kappa)}
$$

where $\theta$ is the angle, $\mu$ is the mean direction (location parameter), $\kappa \geq 0$ is the concentration parameter, and $I_0(\kappa)$ is the modified Bessel function of the first kind of order 0. The von Mises distribution is also known as the circular normal distribution and is the circular analog to the normal distribution. As $\kappa$ increases, the distribution becomes more concentrated around the mean direction $\mu$. When $\kappa = 0$, the distribution reduces to the uniform distribution on the circle. Although we will not dive further into this, the von Mises distribution scales nicely for $p$-dimensional hyperspheres in $\mathbb{R}^p$, making the family of von Mises distributions an attractive and flexible candidate for learning tasks in spherical settings.

\subsection{Variance Encodes Curvature}

In the above parameterization, it is natural to ask how to define the variance in terms of the concentration parameter $\kappa$. Choosing the variance term naively, e.g., a constant, would of course be possible, but would negate the benefits that manifold-learning gives us as a dimensionality reduction technique which preserves local toplogical structure. Thus, we are interested in a principled manner of choosing the variance term that encodes the (dis-)similarity of the topological structure. Intuitively, if we have a directional prior on a region with little topological variation, we may be willing to update our projection more liberally. However, suppose that even a small change in the projection angle moves our data point into an area that is very different from our original projection. In that case, we not be as willing to update our projection.

In manifolds that spherically parameterized, using the inverse absolute Gaussian curvature is a sensible choice to encoding variance. Interestingly, there is an intuitive connection to the score test in likelihood-based inference, which uses the slope at restricted MLE estimates as a measure of closeness/ similarity. Thus, not only does this solution make intuitive topological sense, it has a strong analogue in well-established likelihood-based methods.

The Gaussian curvature of a surface can be defined as the product of the principal curvatures:

$$K = k_1 \cdot k_2$$

where $k_1$ and $k_2$ are the principal curvatures at a point on the surface. For a sphere of radius $R$, we can deduce its Gaussian curvature by considering any point on a sphere of radius $R$. At this point, we can find two perpendicular directions that correspond to the principal curvatures. Due to the perfect symmetry of a sphere, the curvature is the same in all directions at any given point. For a sphere, both principal curvatures are equal to $\frac{1}{R}$ (the reciprocal of the radius). Therefore, for our example, the Gaussian curvature is:

$$K = k_1 \cdot k_2 = \frac{1}{R} \cdot \frac{1}{R} = \frac{1}{R^2}$$

For a unit sphere (where $R = 1$), the Gaussian curvature\footnote{This constant Gaussian curvature of 1 is an intrinsic property of the unit sphere and is related to the fact that the total curvature integrated over the entire sphere equals $4\pi$ (by the Gauss-Bonnet theorem)} is $K = \frac{1}{1^2} = 1$.

\begin{figure}
  \begin{center}
    \includegraphics[width=0.5\textwidth]{../fig/gaussian-curvature.png}
  \end{center}
  \caption{From left to right: a surface of negative Gaussian curvature (hyperboloid), a surface of zero Gaussian curvature (cylinder), and a surface of positive Gaussian curvature (sphere).}\label{fig:gaussian-curvature}
\end{figure}

\subsection{Conjugacy}

A key advantage of using the von Mises distribution are conjugacy results. We lean on the work of \citet{mardia1976}. A von Mises prior distribution with mean direction $\mu$ and concentration parameter $\kappa$ (scaled by $c$) is conjugate to the likelihood. After observing angles $\theta_1,\ldots,\theta_n$, the posterior distribution is proportional to the following expression \citep{mardia1976}:

\[
  f(\mu_i, \mu^* | \{\theta_i\}_{i=1}^n) \propto \text{exp}(\kappa \cdot \sum_{i=1}^n \text{cos}(\theta_i - \mu_i) +\kappa^* \cdot \sum_{i=1}^n \text{cos}(\mu_i - \mu^*))  
\]

where the angles $\{\theta_i\}_{i=1}^n \sim \mathcal{VM}(\mu_i, \kappa)$ and the prior on $\mu_i$ is $\mu_i \sim \mathcal{VM}(\mu^*, \kappa^*).$\footnote{$\mathcal{VM}(\cdot, \cdot)$ denotes the von Mises distribution} In our case, we generated the data with $\mu_i = \mu = \text{circular}(0)$. We can also see that the above form assumes the form of shrinkage estimator\footnote{For the normal case, this procuedure leads to the James-Stein estimator \citep{guttorp1988}.}. The component involving $\{\theta_i\}_{i=1}^n$ is weighted by variance term $\kappa$, in our case the curvature at the projected point. The difference between the mean parameter of the data and the prior mean is weighted by the prior concentration parameter $\kappa^*$. It is interesting to extend this form of a shrinkage estimator to the explicitly topological interpretation we have mentioned above. If the concentration parameter of our data ($\kappa$) is low, we attribute more weight to our "prior" information. The above result is the desired probabilistic framework we sought to establish for this directional prior case.

We are now ready to state the conjugacy result for our 3D spherical example. Given the following von-Mises parameterization of a prior on $\boldsymbol{\theta} \sim \mathcal{VM}(\theta_0, \tau_0 = \frac{1}{\kappa_0})$, we have the posterior mean

\[
  \frac{\tau_0}{\tau_0 + \hat{\tau}} \cdot \theta_0 + \frac{\hat{\tau}}{\tau_0 + \hat{\tau}} \cdot \bar{\theta}
\]

and with posterior variance $(\tau_0 + \hat{\tau})^{-1}$. Here, $\bar{\theta} := \frac{1}{n}\sum_{i=1}^n \theta_i$ and $\hat{\tau} = \frac{1}{\hat{\kappa}}$ where $\hat{\kappa}$ is the observed Gaussian curvature at a point on the manifold. The shrinkage toward the prior mean is immediately evident.

\begin{figure}
  \begin{center}
    \includegraphics[width=1\textwidth]{../fig/posterior-roseplot.png}
  \end{center}
  \caption{Posterior distribution under prior $\theta_0$: Left $\theta_0 = \pi/2$; Right: $\theta_0 = \pi$}\label{fig:posteriors}
\end{figure}

We can see that the posterior mean is essentially the average of the prior and data means. This is because our manifold is a sphere with equivalent curvature at every point. For manifolds with heterogeneous curvature, we would see less reprojection with respect to the prior mean when curvature is more extreme.

\begin{figure}
  \begin{center}

  \end{center}
  \caption{}\label{fig:}
\end{figure}

\section{An Empirical Bayes Extension in the Service of Causality}

Talk about the pre-post treatment idea. What can angles represent? Dont do any code for this, just explain.

Could a prior on $\kappa$ potentially encode anticipated topological changes?

\section{Limitations and Future Direction}

\newpage

\appendix
\section*{Appendix A.} \label{sc:app_a}
In this appendix we prove the following theorem from
Section \ref{sc:intro}

\newpage 
\bibliography{references}

\end{document}

### 2025-04-03 _archive/circular-package.R

@@ -1,0 +1,144 @@
+# ==============================================================================
+# Title: Circular Package
+# Author: Daniel Posmik
+# E-Mail: daniel_posmik@brown.edu
+# ==============================================================================
+
+# Set-Up =======================================================================
+library(dplyr)
+library(tidyverse)
+library(circular)
+
+# Simulation ===================================================================
+# Set random seed for reproducibility
+set.seed(1)
+
+# Parameters
+n <- 500
+mu <- circular(pi, units = "radians")
+rho_values <- c(0.3, 0.6, 0.9)
+colors <- c("blue", "red", "green")
+
+# Create an empty data frame to store results
+results <- data.frame(
+  angle = numeric(0),
+  rho = numeric(0)
+)
+
+# Generate random wrapped normal data with different concentrations
+for (i in 1:length(rho_values)) {
+  samples <- circular::rwrappednormal(n, mu = mu, rho = rho_values[i])
+  results <- rbind(results, data.frame(
+    angle = as.numeric(samples),
+    rho = rep(as.character(rho_values[i]), n)
+  ))
+}
+
+# Circular histogram
+par(mfrow = c(2, 2))
+
+# 1. Rose diagram for all concentrations
+plot(circular(results$angle),
+  stack = TRUE, bins = 36,
+  main = "Rose Diagram of Wrapped Normal Distributions",
+  col = colors[as.factor(results$rho)]
+)
+legend("topright",
+  legend = paste("rho =", rho_values),
+  fill = colors, cex = 0.8, bg = "white"
+)
+
+# 2. Individual rose diagrams for each concentration
+for (i in 1:length(rho_values)) {
+  plot(circular(results$angle[results$rho == as.character(rho_values[i])]),
+    stack = TRUE, bins = 36,
+    main = paste("Rose Diagram (rho =", rho_values[i], ")"),
+    col = colors[i]
+  )
+}
+
+# Linear plot with density curves
+par(mfrow = c(1, 1))
+
+# Convert to data frame for ggplot
+df <- data.frame(
+  value = results$angle,
+  concentration = factor(results$rho, levels = as.character(rho_values))
+)
+
+# Create density plot
+ggplot(df, aes(x = value, fill = concentration)) +
+  geom_density(alpha = 0.5) +
[…]

### 2025-04-03 _archive/ellipsis-projection.R

@@ -1,0 +1,331 @@
+#===============================================================================
+# Title: Prior Information in Manifold Projections
+# Author: Daniel Posmik
+# E-Mail: daniel_posmik@brown.edu
+#===============================================================================
+
+# Set-Up =======================================================================
+library(dplyr)
+library(plotly)
+
+# Simulation ===================================================================
+# Set random seed for reproducibility
+set.seed(123)
+
+# Parameters for the elliptic curve
+a <- 3  # semi-major axis
+b <- 1  # semi-minor axis
+
+# Generate 5 points on an elliptic curve
+n <- 5
+theta <- seq(0, 2*pi, length.out = n+1)[1:n]  # equally spaced angles
+
+# Generate points on the perfect ellipse
+x_ellipse <- a * cos(theta)
+y_ellipse <- b * sin(theta)
+
+# Add Gaussian noise N(0,1) to create observed points
+noise_sd <- 0.2
+x_observed <- x_ellipse + rnorm(n, 0, noise_sd)
+y_observed <- y_ellipse + rnorm(n, 0, noise_sd)
+
+# Generate uniform(0,1) values for the z coordinate
+z_values <- runif(n)
+
+# Create the data frame
+data <- data.frame(
+  x = x_observed,
+  y = y_observed,
+  z = z_values
+)
+
+# Projection ===================================================================
+# Project a point onto the ellipse and return coordinates and angles
+project_to_ellipse <- function(x, y, z, a, b) {
+  # Starting guess for theta (x-y plane angle)
+  theta_guess <- atan2(y/b, x/a)
+  
+  # Function to minimize: squared distance from (x,y,z) to a point on the ellipse
+  distance_function <- function(theta) {
+    ellipse_x <- a * cos(theta)
+    ellipse_y <- b * sin(theta)
+    ellipse_z <- 0 # Since our ellipse only lives in x,y space
+    return(sqrt((x - ellipse_x)^2 + (y - ellipse_y)^2 + (z - ellipse_z)^2))
+  }
+  
+  # Optimize to find the best theta
+  result <- optimize(distance_function, c(0, 2*pi))
+  theta_optimal <- result$minimum
+  
+  # Calculate projected coordinates
+  x_proj <- a * cos(theta_optimal)
+  y_proj <- b * sin(theta_optimal)
[…]

### 2025-04-03 code/01_simulation.R

@@ -1,0 +1,89 @@
+# ==============================================================================
+# Title: Simulation of Data
+# Author: Daniel Posmik
+# E-Mail: daniel_posmik@brown.edu
+# ==============================================================================
+
+# Set-Up =======================================================================
+library(dplyr)
+library(tidyverse)
+library(circular)
+library(plotly)
+
+# Set working directory
+setwd("~/Documents/brown/classes/spring25/php2530-bayes/directional-priors")
+
+# Set random seed for reproducibility
+set.seed(1)
+
+# Generate True Data ===========================================================
+# Sample size
+n <- 500
+
+# Sample from circular normal distribution (wrapped normal)
+theta <- rwrappednormal(n, mu = circular(0), rho = 0.8)
+phi <- rwrappednormal(n, mu = circular(pi/2), rho = 0.8)
+
+# Convert spherical coordinates to Cartesian
+X <- sin(phi) * cos(theta)
+Y <- sin(phi) * sin(theta)
+Z <- cos(phi)
+
+# Create data frame
+df <- data.frame(X, Y, Z)
+
+# Simulate Data ================================================================
+# Noise parameters 
+E <- 0.25
+
+# Add noisy versions of each point
+# X_noisy, Y_noisy, Z_noisy are sampled from N(original point, E)
+df$X_noisy <- df$X + rnorm(n, mean = 0, sd = E)
+df$Y_noisy <- df$Y + rnorm(n, mean = 0, sd = E)
+df$Z_noisy <- df$Z + rnorm(n, mean = 0, sd = E)
+
+# Save data
+write.csv(df, file = "data/01_sim-results.csv")
+
+# Visualization on unit sphere
+sphere_points <- 20
+theta_sphere <- seq(0, 2*pi, length.out = sphere_points)
+phi_sphere <- seq(0, pi, length.out = sphere_points)
+theta_matrix <- matrix(rep(theta_sphere, sphere_points), sphere_points, sphere_points)
+phi_matrix <- matrix(rep(phi_sphere, each = sphere_points), sphere_points, sphere_points)
+x_sphere <- sin(phi_matrix) * cos(theta_matrix)
+y_sphere <- sin(phi_matrix) * sin(theta_matrix)
+z_sphere <- cos(phi_matrix)
+
+# 3D plotly visualization
+fig <- plot_ly() %>%
+  # Add unit sphere
+  add_surface(x = x_sphere, y = y_sphere, z = z_sphere, 
[…]

### 2025-04-03 code/02_projection.R

@@ -1,0 +1,231 @@
+# ==============================================================================
+# Title: Projection of Points
+# Original Author: Daniel Posmik
+# E-Mail: daniel_posmik@brown.edu
+# ==============================================================================
+
+# Set-Up =======================================================================
+library(dplyr)
+library(tidyverse)
+library(circular)
+library(plotly)
+
+# Set working directory
+setwd("~/Documents/brown/classes/spring25/php2530-bayes/directional-priors")
+
+# Import data
+df <- read.csv("data/01_sim-results.csv")
+
+# Projecting data on unit sphere ===============================================
+# Function to compute distance to unit sphere
+sphere_dist <- function(p, noisy_point) {
+  # p is normalized during optimization
+  p_norm <- p / sqrt(sum(p^2))
+  # Compute Euclidean distance
+  return(sqrt(sum((noisy_point - p_norm)^2)))
+}
+
+# Project points to unit sphere
+project_to_sphere <- function(df) {
+  results <- data.frame(
+    X_proj = numeric(nrow(df)),
+    Y_proj = numeric(nrow(df)),
+    Z_proj = numeric(nrow(df)),
+    distance = numeric(nrow(df)),
+    theta = numeric(nrow(df)),
+    phi = numeric(nrow(df))
+  )
+  
+  for (i in 1:nrow(df)) {
+    noisy_point <- c(df$X_noisy[i], df$Y_noisy[i], df$Z_noisy[i])
+    
+    # Use noisy point as initial guess
+    opt <- optim(
+      par = noisy_point,
+      fn = sphere_dist,
+      noisy_point = noisy_point,
+      method = "BFGS"
+    )
+    
+    # Normalize to ensure exactly on sphere
+    proj_point <- opt$par / sqrt(sum(opt$par^2))
+    results$X_proj[i] <- proj_point[1]
+    results$Y_proj[i] <- proj_point[2]
+    results$Z_proj[i] <- proj_point[3]
+    
+    # Calculate distance
+    results$distance[i] <- sqrt(sum((noisy_point - proj_point)^2))
+    
+    # Calculate projection vector (from noisy to projected)
+    proj_vector <- proj_point - noisy_point
+    
[…]

### 2025-04-03 code/03_inference.R

@@ -1,0 +1,246 @@
+# ==============================================================================
+# Title: Bayesian Inference with Spherical Normal Distribution
+# ==============================================================================
+library(dplyr)
+library(tidyverse)
+library(circular)
+library(plotly)
+library(mvtnorm)
+
+# Set working directory
+setwd("~/Documents/brown/classes/spring25/php2530-bayes/directional-priors")
+
+# Import data
+df_proj <- read.csv("data/02_proj-results.csv")
+
+# Posterior inference ==========================================================
+# Function to calculate posterior quantities
+calculate_posterior <- function(prior_theta, prior_kappa = 1) {
+  prior_var <- 1/(prior_kappa)^2
+  data_kappa <- mean(df_proj$kappa)
+  data_var <- 1/(data_kappa)^2
+  data_theta <- mean(df_proj$theta_global) + pi
+  
+  # Posterior estimates
+  post_theta <- (prior_var)/(prior_var + data_var) * prior_theta + 
+    (data_var)/(prior_var + data_var) * data_theta + pi
+  post_var <- 1/(1/prior_var + 1/data_var)
+  
+  # Ensure post_theta is within [0, 2π]
+  post_theta <- post_theta %% (2*pi)
+  
+  return(list(
+    post_theta = post_theta,
+    post_var = post_var
+  ))
+}
+
+# Prior means to visualize
+prior_means <- c(pi/2, pi, 3*pi/2)
+prior_results <- lapply(prior_means, calculate_posterior)
+
+# Create Rose Plot Style Visualization =========================================
+create_rose_plot_style <- function(prior_theta, posterior_result, data_angles, show_legend = FALSE) {
+  # Number of bins for the rose plot
+  n_bins <- 24
+  bin_width <- 2*pi/n_bins
+  
+  # Create bins for data
+  breaks <- seq(0, 2*pi, length.out = n_bins + 1)
+  bin_mids <- breaks[-1] - bin_width/2
+  
+  # Count data in each bin
+  data_bins <- table(cut(data_angles %% (2*pi), breaks = breaks, include.lowest = TRUE))
+  data_bins <- as.numeric(data_bins)
+  
+  # Generate wrapped normal samples for posterior
+  n_samples <- 10000
+  wrapped_samples <- rnorm(n_samples, 
+                           mean = posterior_result$post_theta,
[…]

### 2025-04-10 output/paper.tex
finished first draft with sim included

@@
 \usepackage{subcaption}
+\definecolor{darkorange}{RGB}{204,85,0}
 
@@
 
-\title{	Directional Priors in Manifold Learning \\
+\title{	Directional Priors in Manifold Learning: A Causal Perspective \\
 \vspace{.1in}
@@
 
-\begin{figure}
+\begin{figure}[h!]
   \begin{center}
@@
 Within the principal manifolds framework \citep{Meng2021} -- a replicable and flexible framework for manifold learning -- the process of fitting a manifold to our data contains multiple steps.
-The key idea is that we fit a $d$-dimensional manifold to our $D$-dimensional by minimizing the sum of squares between our data and the proposed manifold.
+The key idea of the fitting step is that we fit a $d$-dimensional manifold to our $D$-dimensional by minimizing the sum of squares between our data and the proposed manifold.
 An important extension to linear dimensionality reduction, i.e.
@@
 We note that -- although certainly intuitive -- this topological structure is not only limited to spatial abstraction, but may be extended to arbitrary dimensions of interest.
-This framework was pioneered as an extension to the PCA algorithm with curves (\citet{HastieStuetzle1989}, \citet{Tibshirani1992}) and has since found a myriad of applications in higher-dimensional extensions.
+This framework was pioneered as an extension to the PCA algorithm with curves (\citet{HastieStuetzle1989}, \citet{Tibshirani1992}) and has since found a myriad of applications in higher-dimensional settings.
 
@@
 
-\section{Background}
+\section{Background}\label{sc:background}
 
@@
 
-where we consider each projection function $f$ to be a member of an arbitrary function space $\mathcal{F}$. We define the distance metric as the $L^2$ distance.
+where we consider each projection function $f$ to be a member of an arbitrary Sobolev space $\mathcal{F}$. We define the distance metric as the $L^2$ distance.
 
-If there exists only one projection point $f(x_i)$ for every $x_i$, every $f \in \mathcal{F}$ is one-to-one and onto ("bijective") mapping.
+If there exists only one projection point $f(x_i)$ for every $x_i$, every $f \in \mathcal{F}$ is a one-to-one and onto ("bijective") mapping.
 We find it interesting to highlight that the projection functions in the PCA algorithm are inherently bijective, and for inferential purposes, this is a property that is often taken for granted\footnote{This is because a principal axes is a straight line, i.e.
@@
 
-\section{Directional Priors}
+\section{Directional Priors}\label{sc:directional-priors}
 
@@
 
-\subsection{Formulating Directional Priors}
+\subsection{Formulating Directional Priors}\label{sc:dir-priors}
 
-Formulating a directional prior is not trivial.
-For one, it is important to set a reference point for both the directional prior and the angles.
-It is important that all angles are measured from the same reference point.
-For simplicity, we choose the origin.
-Now, choosing a directional prior in some direction $\mu \in [0, \pi]$ allows\footnote{Note that we choose prior means are only meaningful in $[0, \pi]$ This is because for any prior mean greater than $\pi$, we would consider the closer angle in the opposite direction.} us to incorporate information on where the projected points should be concentrated on the manifold.
-We only posit a prior on the random set $\{\theta_i\}_{i=1}^n$, expressing some prior distributional belief on the concentrations of points in the XY-plane.
-In this example, we consider the angles $\{\phi_i\}_{i=1}^n$ to be fixed, making our example univariate.
-An extension to the multivariate case would be straightforward by using the multivariate analogoues of our likelihood and prior distribution functions.
+Formulating a directional prior requires careful consideration of reference points.
+All angles must be measured from the same reference point—for simplicity, we choose the origin.
+A directional prior with mean direction $\mu \in [0, \pi]$ allows us to incorporate information about where projected points should concentrate on the manifold.
+We only posit a prior on the random set $\{\theta_i\}_{i=1}^n$, expressing prior distributional beliefs on point concentrations in the XY-plane.
+For this example, we consider angles $\{\phi_i\}_{i=1}^n$ fixed, making our example univariate.
+Extension to the multivariate case is straightforward using multivariate analogs of our likelihood and prior distributions.
 
-Our posterior distribution on the random vector $\theta$ can be expressed as follows.
+Our posterior distribution on the random vector $\theta$ can be expressed as:
 
-\[
+$$
 \begin{aligned}
@@
 \end{aligned}
-\]
+$$
 
-where $\boldsymbol{\mu}$ is our mean vector that we posit a prior on, i.e.
-$\mu \sim f_{\mu}(\mu)$.
-To form our likelihood, we rely on the wrapped ("spherical") normal distribution.
-The density of a wrapped normal is given by
+where $\mu$ is our mean vector with prior $\mu \sim f_{\mu}(\mu)$. For our likelihood, we initially consider the wrapped normal distribution with density:
 
@@
 
-with the infinite support reflected in the $\mathit{mod}(k)$ argument.
-Integrating over $2\pi$ is equivalent to integrating over $\mathbb{R}$ in the canonical parameterization.
-Now, a disadvantage of using the above form is the infinite series.
-Deriving moments is not straightforward and requires (inverse) Fourier transforms.
-Thus, we will use the von Mises distribution, an analogue to the spherical normal that is easier to handle.
-The density of a von Mises distribution is given by
+However, this infinite series presents computational challenges. Instead, we use the von Mises distribution, which provides a more tractable circular analog to the normal distribution:
 
@@
 
-where $\theta$ is the angle, $\mu$ is the mean direction (location parameter), $\kappa \geq 0$ is the concentration parameter, and $I_0(\kappa)$ is the modified Bessel function of the first kind of order 0.
-The von Mises distribution is also known as the circular normal distribution and is the circular analog to the normal distribution.
-As $\kappa$ increases, the distribution becomes more concentrated around the mean direction $\mu$.
-When $\kappa = 0$, the distribution reduces to the uniform distribution on the circle.
-Although we will not dive further into this, the von Mises distribution scales nicely for $p$-dimensional hyperspheres in $\mathbb{R}^p$, making the family of von Mises distributions an attractive and flexible candidate for learning tasks in spherical settings.
+where $\theta$ is the angle, $\mu$ is the mean direction, $\kappa \geq 0$ is the concentration parameter (analogous to precision, or inverse variance), and $I_0(\kappa)$ is the modified Bessel function of first kind of order 0.
+As $\kappa$ increases, the distribution concentrates more tightly around $\mu$.
+When $\kappa = 0$, the distribution becomes uniform on the circle.
+The von Mises distribution scales naturally for hyperspheres in $\mathbb{R}^p$, making it ideal for spherical settings in manifold learning applications.
 
-\subsection{Variance Encodes Curvature}
+\subsection{Variance Encodes Curvature}\label{sc:curvature}
 
-In the above parameterization, it is natural to ask how to define the variance in terms of the concentration parameter $\kappa$.
-Choosing the variance term naively, e.g., a constant, would of course be possible, but would negate the benefits that manifold-learning gives us as a dimensionality reduction technique which preserves local toplogical structure.
-Thus, we are interested in a principled manner of choosing the variance term that encodes the (dis-)similarity of the topological structure.
-Intuitively, if we have a directional prior on a region with little topological variation, we may be willing to update our projection more liberally.
-However, suppose that even a small change in the projection angle moves our data point into an area that is very different from our original projection.
-In that case, we not be as willing to update our projection.
+For our concentration parameter $\kappa$, naive selection (e.g., a constant) would be possible but would negate the benefits of manifold learning as a topology-preserving dimensionality reduction technique.
+Instead, we propose a principled approach that encodes local topological structure into the variance parameter.
+Intuitively, in regions with little topological variation, we might update our projection more liberally.
+Conversely, if a small change in projection angle moves our data point into a topologically distinct region, we would be more conservative with updates.
 
-In manifolds that spherically parameterized, using the inverse absolute Gaussian curvature is a sensible choice to encoding variance.
-Interestingly, there is an intuitive connection to the score test in likelihood-based inference, which uses the slope at restricted MLE estimates as a measure of closeness/ similarity.
-Thus, not only does this solution make intuitive topological sense, it has a strong analogue in well-established likelihood-based methods.
+In spherically parameterized manifolds, the inverse absolute Gaussian curvature provides a natural variance encoding.
+This approach has an elegant connection to the score test in likelihood-based inference, which uses the slope at restricted MLE estimates as a similarity measure.
+Thus, our solution has both topological intuition and established statistical parallels in traditional inference methods.
 
@@
 where $k_1$ and $k_2$ are the principal curvatures at a point on the surface.
-For a sphere of radius $R$, we can deduce its Gaussian curvature by considering any point on a sphere of radius $R$.
-At this point, we can find two perpendicular directions that correspond to the principal curvatures.
-Due to the perfect symmetry of a sphere, the curvature is the same in all directions at any given point.
-For a sphere, both principal curvatures are equal to $\frac{1}{R}$ (the reciprocal of the radius).
-Therefore, for our example, the Gaussian curvature is:
+For a sphere of radius $R$, due to perfect symmetry, both principal curvatures equal $\frac{1}{R}$, giving a Gaussian curvature of $K = \frac{1}{R^2}$.
+For a unit sphere, the Gaussian curvature is $K = 1$, an intrinsic property related to the total curvature integrated over the entire sphere equaling $4\pi$ by the Gauss-Bonnet theorem.
 
-$$K = k_1 \cdot k_2 = \frac{1}{R} \cdot \frac{1}{R} = \frac{1}{R^2}$$
-
-For a unit sphere (where $R = 1$), the Gaussian curvature\footnote{This constant Gaussian curvature of 1 is an intrinsic property of the unit sphere and is related to the fact that the total curvature integrated over the entire sphere equals $4\pi$ (by the Gauss-Bonnet theorem)} is $K = \frac{1}{1^2} = 1$.
-
-\begin{figure}
+\begin{figure}[h!]
   \begin{center}
@@
 
-\subsection{Conjugacy}
+This curvature-based approach provides a meaningful way to weight observations based on their location in the manifold's topological structure.
+Areas with high curvature (significant bending of the manifold) will contribute less to updating our beliefs than areas with lower curvature (flatter regions), which aligns with our geometric intuition about confidence in projections.
 
-A key advantage of using the von Mises distribution are conjugacy results.
-We lean on the work of \citet{mardia1976}.
-A von Mises prior distribution with mean direction $\mu$ and concentration parameter $\kappa$ (scaled by $c$) is conjugate to the likelihood.
-After observing angles $\theta_1,\ldots,\theta_n$, the posterior distribution is proportional to the following expression \citep{mardia1976}:
+\subsection{Conjugacy Results}\label{sc:conjugacy}
 
-\[
-  f(\mu_i, \mu^* | \{\theta_i\}_{i=1}^n) \propto \text{exp}(\kappa \cdot \sum_{i=1}^n \text{cos}(\theta_i - \mu_i) +\kappa^* \cdot \sum_{i=1}^n \text{cos}(\mu_i - \mu^*))  
-\]
+A key advantage of the von Mises distribution is conjugacy, which allows for straightforward Bayesian updates of our directional priors.
+Following the work of \citet{mardia1976}, a von Mises prior with mean direction $\mu$ and concentration parameter $\kappa$ is conjugate to the likelihood.
+After observing angles $\theta_1,\ldots,\theta_n$, the posterior distribution takes the form:
 
-where the angles $\{\theta_i\}_{i=1}^n \sim \mathcal{VM}(\mu_i, \kappa)$ and the prior on $\mu_i$ is $\mu_i \sim \mathcal{VM}(\mu^*, \kappa^*).$\footnote{$\mathcal{VM}(\cdot, \cdot)$ denotes the von Mises distribution} In our case, we generated the data with $\mu_i = \mu = \text{circular}(0)$.
-We can also see that the above form assumes the form of shrinkage estimator\footnote{For the normal case, this procuedure leads to the James-Stein estimator \citep{guttorp1988}.}.
-The component involving $\{\theta_i\}_{i=1}^n$ is weighted by variance term $\kappa$, in our case the curvature at the projected point.
-The difference between the mean parameter of the data and the prior mean is weighted by the prior concentration parameter $\kappa^*$.
-It is interesting to extend this form of a shrinkage estimator to the explicitly topological interpretation we have mentioned above.
-If the concentration parameter of our data ($\kappa$) is low, we attribute more weight to our "prior" information.
-The above result is the desired probabilistic framework we sought to establish for this directional prior case.
+$$
+f(\mu_i, \mu^* | \{\theta_i\}_{i=1}^n) \propto \exp(\kappa \cdot \sum_{i=1}^n \cos(\theta_i - \mu_i) +\kappa^* \cdot \sum_{i=1}^n \cos(\mu_i - \mu^*))
+$$
 
-We are now ready to state the conjugacy result for our 3D spherical example.
-Given the following von-Mises parameterization of a prior on $\boldsymbol{\theta} \sim \mathcal{VM}(\theta_0, \tau_0 = \frac{1}{\kappa_0})$, we have the posterior mean
+where angles $\{\theta_i\}_{i=1}^n \sim \mathcal{VM}(\mu_i, \kappa)$ and the prior is $\mu_i \sim \mathcal{VM}(\mu^*, \kappa^*)$.
+In our case, the data were generated with $\mu_i = \mu = \text{circular}(0)$.
 
-\[
-  \frac{\tau_0}{\tau_0 + \hat{\tau}} \cdot \theta_0 + \frac{\hat{\tau}}{\tau_0 + \hat{\tau}} \cdot \bar{\theta}
-\]
+The posterior form reveals a shrinkage estimator structure.
+The component involving observed angles $\{\theta_i\}_{i=1}^n$ is weighted by $\kappa$, which in our framework represents the curvature at the projected point.
+The difference between the data mean and prior mean is weighted by the prior concentration parameter $\kappa^*$.
+This structure extends shrinkage estimation to our topological interpretation: when the data concentration parameter $\kappa$ is low, more weight is attributed to our prior information.
 
-and with posterior variance $(\tau_0 + \hat{\tau})^{-1}$.
-Here, $\bar{\theta} := \frac{1}{n}\sum_{i=1}^n \theta_i$ and $\hat{\tau} = \frac{1}{\hat{\kappa}}$ where $\hat{\kappa}$ is the observed Gaussian curvature at a point on the manifold.
-The shrinkage toward the prior mean is immediately evident.
+For our 3D spherical example, given a prior $\boldsymbol{\theta} \sim \mathcal{VM}(\theta_0, \tau_0 = \frac{1}{\kappa_0})$, the posterior mean is:
 
-\begin{figure}
+$$
+\underbrace{\frac{\tau_0}{\tau_0 + \hat{\tau}} \cdot \theta_0}_{\text{Weighted Prior Vector}} + \underbrace{\frac{\hat{\tau}}{\tau_0 + \hat{\tau}} \cdot \bar{\theta}}_{\text{Weighted Data Vector}}
+$$
+
+with posterior variance $(\tau_0 + \hat{\tau})^{-1}$, where $\bar{\theta} := \frac{1}{n}\sum_{i=1}^n \theta_i$ and $\hat{\tau} = \frac{1}{\hat{\kappa}}$ with $\hat{\kappa}$ being the observed Gaussian curvature at the manifold point.
+The shrinkage toward the prior mean is immediately evident in this expression.
+
+\begin{figure}[h!]
   \begin{center}
@@
 
-We can see that the posterior mean is essentially the average of the prior and data means.
-This is because our manifold is a sphere with equivalent curvature at every point.
-For manifolds with heterogeneous curvature, we would see less reprojection with respect to the prior mean when curvature is more extreme.
+Figure \ref{fig:posteriors} demonstrates that the posterior mean effectively balances the prior and data means.
+This occurs because our manifold is a sphere with equivalent curvature at every point.
+For manifolds with heterogeneous curvature, we would observe less reprojection with respect to the prior mean in regions where curvature is more extreme, thus preserving the topological characteristics of the data.
+
+The conjugacy relationship for von Mises distributions has an elegant geometric interpretation.
+For a von Mises likelihood $f(x|\mu) \propto \exp(\kappa \cos(x-\mu))$ and a von Mises prior on the mean direction $\pi(\mu) \propto \exp(\kappa_0 \cos(\mu-\mu_0))$, the posterior distribution is also von Mises: $\pi(\mu|x) \propto \exp(\kappa_n \cos(\mu-\mu_n))$.
+The posterior parameters can be expressed in terms of resultant vectors:
+
+$$\kappa_n\cos(\mu_n) = \kappa_0\cos(\mu_0) + \kappa\cos(x)$$
+$$\kappa_n\sin(\mu_n) = \kappa_0\sin(\mu_0) + \kappa\sin(x)$$
+
+These equations can be rewritten in vector form as:
+
+$$\kappa_n = \sqrt{(\kappa_0\cos(\mu_0) + \kappa\cos(x))^2 + (\kappa_0\sin(\mu_0) + \kappa\sin(x))^2}$$
+
+and
+
+$$\mu_n = \text{atan2}(\kappa_0\sin(\mu_0) + \kappa\sin(x), \kappa_0\cos(\mu_0) + \kappa\cos(x))$$
+
+This representation yields a powerful geometric interpretation: the posterior mean direction and concentration are determined by the vector sum of the prior and likelihood information vectors in $\mathbb{R}^2$.
+Each vector has direction given by unit vectors pointing at angles $\mu_0$ and $x$ respectively, with magnitude determined by concentration parameters $\kappa_0$ and $\kappa$.
+The resultant vector's length determines the posterior concentration parameter $\kappa_n$, while its direction determines the posterior mean direction $\mu_n$.
+
+This insight demonstrates that when prior and likelihood point in similar directions, the posterior becomes more concentrated (higher $\kappa_n$), reflecting increased certainty.
+Conversely, conflicting directions lead to a less concentrated posterior (lower $\kappa_n$), reflecting uncertainty.
+
+The posterior mean naturally represents a weighted compromise between prior and data, with weights proportional to their respective concentration parameters.
+For multiple observations, the vectors from each observation simply add to the prior vector, providing an efficient sequential updating mechanism.
+This vector-based interpretation elegantly connects Bayesian inference on the circle with vector addition in the plane, making it particularly suitable for our manifold learning framework where directional data naturally arises from projections onto curved surfaces.
+
+\section{A Causal Extension -- Toward a Manifold RD Estimator}\label{sc:causal}
+
+\textcolor{red}{Make it crystal clear that the posterior variance is the Bayesian thing here!!}
+
+An interesting application of directional priors is a causal setting where we seek to estimate treatment effects after some treatment has been administered.
+A popular class of estimators for comparing changes in outcomes of interest close to a treatment cut-off are regression discontinuity (RD) estimators.
+RD estimators exploit the the balance between units close to the treatment cut-off, assuming that treatment does not depend on any unit characteristics.
+Recovering causal effects in RD estimators hinges on the assumption that in terms of unit characteristics, treatment assignment is quasi-random very close to the cut-off.
+Popular examples in literature include measuring the effect of advanced coursework on students above some cutoff, comparing their outcomes to students just below said cutoff.
+
+In the principal manifold learning setup, directional priors offer a promising Bayesian approach to causal inference.
+For simplicity, assume that treatment is administered at time point $t^*$.
+Then, all units in $(-\infty, t^*)$ are pre-treatment units, and all units in $[t^*, \infty)$ are post-treatment units.
+In a canonical RD setting, under a set of identification assumptions, we can recover an average treatment by considering values of $Y_i$ immediately to the right (say $Y_{it} |_{t^* + \delta}$) and left ($Y_{it} |_{t^* - \delta}$) of the cutoff.
+This yields the causal estimate
+
+$$
+\hat{\tau}= \bar{Y}|_{t = t^* + \delta} - \bar{Y}|_{t = t^* - \delta} 
+$$
+
+The above estimator is commonly measured in the Local Linear Regression (LLR) framework, i.e.
+measure the ATE as the difference in post- and pre-treatment means, given that they are sufficiently close to the treatment cutoff $t^*$.
+\citet{Branson2019} considers a Bayesian RD setting where a Gaussian process prior is placed on the pre- and post-treatment mean response functions.
+We believe this work, although not directly related to the use of topological information in causal inference, encapsulates the desire for more flexible, data-driven, and principled causal inference through Bayesian paradigms.
+
+\subsection{Intuition}
+
+In a principal manifolds framework, we can also exploit this notion of quasi-randomness to recover causal effects.
+Specifically, we propose the estimation of causal treatment effects using the posterior density of projection angles with empirical prior variances encoding the local topological structure of pre-treatment manifolds.
+For this procedure, suppose we use the set of pre-treatment units $\{Y_{it}\}_{i=1}^n|_{t < t^*}$ to fit the $d$-dimensional manifold $\mathcal{M}_d^{t < t^*}$.
+As we have shown in Section~\ref{sc:directional-priors}, a polar reparameterization allows us to recover the projection angles $\{\theta_i\}$ as well as their mean which we shall denote as $\bar{\theta}^{t < t^*}$.
+This mean pre-treatment projection angle $\bar{\theta}^{t < t^*}$ has a curvature which we will denote as $\kappa^{t<t^*}$.
+Together, the mean pre-treatment projection angle and curvature sufficiently parameterize the first two moments of the pre-treatment outcome of interest.
+
+After treatment happens at $t=t^*$, we can fit the post-treatment manifold $\mathcal{M}_d^{t \geq t^*}$.
+In the spirit of LLR-based causal estimation, we want to emphasize that $\mathcal{M}_d^{t \geq t^*}$ is *not* fit iteratively for each post-treatment unit, i.e.
+we do not obtain the post-treatment unit by combining the post-treatment units with pre-treatment units.
+Instead, our pre- and post-treatment manifolds are topological analogues to $\bar{Y}|_{t = t^* - \delta}$ and $\bar{Y}|_{t = t^* + \delta}$, respectively.
+That being said, it may be interesting to let the choice of manifold (e.g., the shape, highest order, etc.) from the pre-treatment case inform the post-treatment case, effectively borrowing prior topological information on a more global level.
+\footnote{This is a really interesting direction in the Bayesian paradigm and touches on the notion of persistent homology.
+Although this discussion goes beyond the scope of this paper, it highlights how nuanced various levels/ sources of topological information can be.} Now, suppose that we observe the outcomes of $m$ additional units in the immediate post-treatment period, then the data $\{Y_{it}\}_{i=n+1}^{m}|_{t \geq t^*}$ allows us to fit the manifold $\mathcal{M}_d^{t \geq t^*}$.
+The mean projection angle of $\mathcal{M}_d^{t \geq t^*}$, i.e.
+$\bar{\theta}^{t \geq t^*}$ will update from $\bar{\theta}^{t < t^*}$, and yield a new posterior estimate for the projection angle.
 
@@
   \begin{center}
+\begin{tikzpicture}[scale=0.9]
+    % Define colors
+    \definecolor{darkgrey}{RGB}{80,80,80}
+    \definecolor{darkorange}{RGB}{204,85,0}
 
+    % Time axis
+    \draw[->, thick] (-0.5,0) -- (10.5,0) node[right] {$t$};
+
+    % Pre-treatment and post-treatment labels
+    \node[align=center, text width=4cm] at (2.5,-0.5) {\text{Pre-treatment}};
+    \node[align=center, text width=4cm] at (7.5,-0.5) {\text{Post-treatment}};
+
+    % Treatment cutoff vertical line
+    \draw[dashed, thick] (5,0) -- (5,7);
+    \node at (5,-0.5) {$t^*$};
+
+    % Pre-treatment manifold as a regular ellipse (dark grey)
+    \draw[thick, darkgrey] (2.5,3) ellipse (1.5 and 2);
+    \node[darkgrey] at (2.5,5.7) {$\mathcal{M}_d^{t < t^*}$};
+
+    % Pre-treatment projection angle (dark grey)
+    \draw[->, thick, darkgrey] (2.5,3) -- (3.5,4.4);
+    \node[darkgrey] at (4.2, 4.5) {$\bar{\theta}^{t < t^*}$};
+
+    % Post-treatment: ghost of pre-treatment manifold (dashed)
+    \draw[thick, dashed, darkgrey] (7.5,3) ellipse (1.5 and 2);
+
+    % Post-treatment manifold as a pointier ellipse, rotated 12 degrees (darkorange)
+    \begin{scope}[shift={(7.5,3)}, rotate=12]
+        \draw[thick, darkorange] (0,0) ellipse (1.2 and 2.2);
+    \end{scope}
+    \node[darkorange] at (7.5,5.7) {$\mathcal{M}_d^{t \geq t^*}$};
+
+    % Two projection angles in post-treatment period
+    \draw[->, thick, darkorange] (7.5,3) -- (7.8,4.85);  % Updated angle (darkorange)
+    \node[darkorange] at (8.5, 5.2) {$\bar{\theta}^{t \geq t^*}$};
+    \draw[->, thick, darkgrey] (7.5,3) -- (8.6,4.4);  % Original angle direction (dashed, dark grey)
+     \node[darkgrey] at (9.4, 4.5) {$\bar{\theta}^{t < t^*}$};
+
+    % Data points on pre-treatment manifold (dark grey)
+    \foreach \x/\y in {1.5/3.5, 2/2, 3/2, 3.5/3.5, 2.5/4.5, 2.5/1.5} {
+        \fill[darkgrey] (\x,\y) circle (2pt);
+    }
+
+    % Data points on post-treatment manifold (dark grey)
+    \foreach \x/\y in {6.8/3.8, 7/2, 8/2.2, 8.3/3.7, 7.5/4.8, 7.3/1.3, 7/5} {
+        \fill[darkorange] (\x,\y) circle (2pt);
+    }
+
+    % Legend
+    \begin{scope}[shift={(0,-1.5)}]
+        % Rectangle around the legend - made even wider
+        \draw[black] (0,0) rectangle (11,-1.5);
+
+        % Legend title
+        %\node[align=center] at (5.5,-.3) {\textbf{Legend}};
+
+        % Pre-treatment elements (darkgrey)
+        \draw[thick, darkgrey] (1.5,-0.7) -- (2.5,-0.7);
+        \node[darkgrey, align=left, text width=7.5cm] at (6.8,-0.7) {Pre-treatment manifold and projection};
+
+        % Post-treatment elements (darkorange)
+        \draw[thick, darkorange] (1.5,-1.2) -- (2.5,-1.2);
+        \node[darkorange, align=left, text width=7.5cm] at (6.8,-1.2) {Post-treatment manifold and projection};
+    \end{scope} 
+  \end{tikzpicture}
   \end{center}
-  \caption{}\label{fig:}
+  \caption{The Manifold RD Estimator}\label{fig:manifold-rd}
 \end{figure}
 
-\section{An Empirical Bayes Extension in the Service of Causality}
+\subsection{The Estimator}
 
-Talk about the pre-post treatment idea. What can angles represent? Dont do any code for this, just explain.
+Combining our insights from RD estimators with the conjugacy results of Section~\ref{sc:conjugacy}, we propose a Bayesian manifold RD estimator in its simplest form.
+After having the pre- and post-treatment manifold, consider subtracting the pre-treatment distribution of angles from the post-treatment distribution.
+This yields a RD estimator of the form:
 
-Could a prior on $\kappa$ potentially encode anticipated topological changes?
+$$
+f_{\tau | \theta} = 
+\mathcal{VM}
+\left(
+  \bar{\theta}^{t \geq t^*} - \bar{\theta}^{t < t^*}, \sigma^2 
+\right)
+$$
+
+where
+
+\[
+  \sigma^2 = \sqrt{\left[\kappa^{t < t^*}\cos(\bar{\theta}^{t < t^*}) + \textcolor{darkorange}{\kappa^{t \geq t^*}\cos(\bar{\theta}^{t \geq t^*})}\right]^2 
+  + 
+\left[\kappa^{t < t^*}\sin(\bar{\theta}^{t < t^*}) + \textcolor{darkorange}{\kappa^{t \geq t^*}\sin(\bar{\theta}^{t \geq t^*})}\right]^2}
+\]
+
+by the conjugacy results of Section~\ref{sc:conjugacy}.
+
+Here, we estimate the Average Treatment Effect (ATE) as the difference between the mean projection angles of the post- and pre-treatment manifold, respectively.
+As illustrated in Figure~\ref{fig:manifold-rd}, this difference captures the directional shift in data geometry following treatment intervention.
+The posterior variance, expressed as the Euclidean norm of the difference between weighted directional vectors, quantifies our uncertainty about this effect.
+Intuitively, a larger norm indicates greater topological dissimilarity between pre- and post-treatment manifolds, suggesting a more substantial causal impact.
+
+To illustrate this, let us recycle the toy example from Section~\ref{sc:conjugacy}.
+Suppose we have a post-treatment mean projection angle, $\bar{\theta}^{t \geq t^*} = \pi$ and a pre-treatment mean projection angle, $\bar{\theta}^{t < t^*} = \pi/2$.
+We consider both of these manifolds to be circles, but the post-treatment manifold has radius $2$.
+Thus, the manifold we have fit grows from a unit circle (radius = $1$) to a post-treatment circle with radius $2$.
+This means that while the pre-treatment Gaussian curvature stays 1, i.e.
+$\kappa^{t < t^*} = 1$, our post-treatment curvature is now $\kappa^{t \geq t^*} = \frac{1}{R^2} = \frac{1}{4}$.
+We simulate 500 draws and observe the following results:
+
+\begin{figure}[h!]
+  \begin{center}
+    \includegraphics[width=1\textwidth]{../fig/rd-sim.png}
+  \end{center}
+  \caption{Simulation of Treatment Effect $\hat{\tau}$}\label{fig:rd-sim}
+\end{figure}
+
+We can see that the resulting treatment effect estimate $\hat{\tau}$ is indeed very close to $\frac{\pi}{2}$. The variance term is slightly above 1, i.e. greater than the individual parts.
+
+When compared to well-established LLR methods, the mean projection angles are simply a spherical reparameterization of the outcome variable of interest.
+This formulation naturally preserves identification assumptions required for causal interpretation while accommodating the topological abstractions within the principal manifold framework.
+The magnitude of the posterior variance directly reflects how drastically treatment has altered the local geometry of our outcome space, e.g., if we have a large difference, we want to encode that topological uncertainty into our ATE.
+
+An important point of emphasis is what this procedure is not.
+Namely, we are not simply updating our fitted manifold with the new data, and letting the posterior angle update iteratively with each new post-treatment observation.
+Rather, we are using the pre-treatment manifold to borrow topological information for our post-treatment mean vector.
+In that sense, after having fit $\mathcal{M}_d^{t \geq t^*}$ to the post-treatment data only, we assign the prior mean (i.e.
+$\bar{\theta}^{t<t^*}$) a weight of zero, only using the post-treatment observations.
+
+We round out this section with an appealing information-theoretic point of view.
+This manifold RD estimate is fundamentally analogous to the existing methods.
+The main difference is that this procedure accounts for the topological information of our embedding space.
+Existing methods are quick to discard this information and assume it away.
+Principled manifold learning approaches combined with the flexible probabilistic advantages of Bayesian methods can therefore yield estimates that respect both prior and topological information, resulting in more flexible and data-driven inference.
 
 \section{Limitations and Future Direction}
+
+Future work could address the use of priors for the construction of the mean projection vectors in the pre- and post-treatment periods, respectively.
 
@@
 \section*{Appendix A.} \label{sc:app_a}
-In this appendix we prove the following theorem from
-Section \ref{sc:intro}
+All code and documentation can be found here. [Insert Link]
 

### 2025-04-10 output/proposal.qmd
finished first draft with sim included

--- 2025-04-03 output/paper.tex
@@
-\documentclass[twoside,11pt]{article}
+---
+title: "Directional Priors in Manifolds"
+subtitle: "A Causal Interpretation"
+author: "Daniel Posmik"
+date: today
+format: pdf
+toc: false
+number-sections: true
+---
 
-\usepackage{paper}
-\usepackage{tikz}     
-\usepackage{xcolor}
-\usepackage{amsmath}
-\usetikzlibrary{calc}
-\usepackage{graphicx}
-\usepackage{subcaption}
+# Note
 
-\ShortHeadings{
-%% short title here
-Directional Priors in Manifolds
-}{
-% last name
-Posmik}
-\firstpageno{1}
+Prof.
+Oganisian -- as discussed, I have attached the abstract and the introduction of my paper.
+It already exists in its rough form, but I have not yet written the causal section.
+Please let me know what suggestions and thoughts you have, I would be eager to incorporate them.
 
-\begin{document}
+# Abstract
 
-\title{	Directional Priors in Manifold Learning \\
-\vspace{.1in}
-PHP2530			
-}
+This paper introduces an approach for incorporating directional prior information into manifold learning, specifically within the principal manifolds framework.
+We formulate a Bayesian method that leverages directional statistics to update prior knowledge about the orientation of data points projected onto manifolds.
+With a spherical example in $\mathbb{R}^3$, we demonstrate how directional uncertainty can be modeled using the von Mises distributions.
+We exploit established conjugacy results, letting the concentration parameter encode the (fitted) local Gaussian curvature of the manifold at a projection point.
+Within this (empirical) Bayesian paradigm, we also explore the Manifold analogue of a causal regression discontinuity estimator where pre- and post treatment manifolds are fitted.
+Treating the curvature of pre-treatment manifolds as prior variance information, we explore the construction of a causal estimator as the posterior projection angles change in the post-treatment period.
+This paper outlines the foundation for incorporating prior directional knowledge into the Principal Manifolds framework, suggesting applications in fields like cancer medicine, and exploring the viability of causal effect identification within a Bayesian paradigm.
 
-\author{ Daniel C. Posmik }
-
-\maketitle
-\date{4 }
-
-\section{Introduction} \label{sc:intro}
+# Introduction
 
 Manifold learning is dimensionality reduction technique that has proven useful in settings where data are high-dimensional and non-linear.
-Often, manifold learning algorithms are used when the topological structure of the data are to be preserved in a statistical learning task\footnote{For an introduction, see \citet{Meila2023}}.
+Often, manifold learning algorithms are used when the topological structure of the data are to be preserved in a statistical learning ^[For an introduction, see (Meila2023)].
 Within the manifold learning framework, data are assumed to live on a lower dimensional manifold and are corrupted by high-dimensional noise.
-We say that $D$-dimensional data can be embedded in $d$-dimensional manifold where $d \leq D$ but generally $d \ll D$ .
+We say that $D$-dimensional data can be embedded in $d$-dimensional manifold where $d \leq D$ but generally $d \ll D$.
 
-\begin{figure}
-  \begin{center}
-    \begin{tikzpicture}[scale=0.85]
-      % Define styles
-      \tikzset{
-        manifold/.style={blue, thick, fill=blue!10},
-        noise/.style={gray, thick},
-        observed/.style={blue, thick, fill=blue!10, opacity=0.7},
-        arrow/.style={->, >=stealth, thick},
-        label/.style={font=\normalfont\normalsize}
-      }
-      % Low-dimensional manifold (left)
-      \begin{scope}[xshift=-5cm]
-        \draw[manifold] plot[smooth, tension=0.8] coordinates {(-1.5,0) (-0.5,1) (1,1.2) (1.5,0) (1,-1) (-0.5,-1) (-1.5,0)};
+![Manifold Learning](../fig/manifold-learning.png){#fig-noisy-data width=75%}
 
-        % Draw some points on the manifold
-        \node[circle, draw, inner sep=1pt, fill=blue!50] at (-0.7,0.3) {$f$};
-        \node[circle, draw, inner sep=1pt, fill=blue!50] at (0.8,0.5) {$f$};
-        \node[circle, draw, inner sep=1pt, fill=blue!50] at (0.2,-0.6) {$f$};
-      \end{scope}
-      % High-dimensional noise (center)
-      \begin{scope}
-        % Draw noise as radial lines from center point
-        \fill (0,0) circle (0.1);
-        \foreach \i in {0,10,...,350} {
-          \draw[noise] (0,0) -- (\i:1.2+0.3*rnd);
-        }
-      \end{scope}
-      % Observed data (right)
-      \begin{scope}[xshift=5cm]
-        % First draw the noise pattern
-        \fill (0,0) circle (0.1);
-        \foreach \i in {0,10,...,350} {
-          \draw[noise] (0,0) -- (\i:1.2+0.3*rnd);
-        }
-
-        % Then overlay the manifold with transparency
-        \draw[observed] plot[smooth, tension=0.8] coordinates {(-1.5,0) (-0.5,1) (1,1.2) (1.5,0) (1,-1) (-0.5,-1) (-1.5,0)};
-
-        % Draw some points on the manifold
-        \node[circle, draw, inner sep=1pt, fill=blue!50, opacity=0.7] at (-0.7,0.3) {$f$};
-        \node[circle, draw, inner sep=1pt, fill=blue!50, opacity=0.7] at (0.8,0.5) {$f$};
-        \node[circle, draw, inner sep=1pt, fill=blue!50, opacity=0.7] at (0.2,-0.6) {$f$};
-      \end{scope}
-      % Draw the operation symbols
-      \node at (-2.5,0) {$+$};
-      \draw[arrow] (2,0.5) -- (3,0.5);
-      \draw[arrow] (3,-0.5) -- (2,-0.5);
-      % Draw the curved arrows for parametrization and embedding
-      \draw[arrow, green!50!black, thick] (-4,2.5) to[bend left=30] node[above, font=\normalfont\normalsize] {Parameterization} (4,2.5);
-      \draw[arrow, orange, thick] (4,-2.5) to[bend left=30] node[below, font=\normalfont\normalsize] {Embedding} (-4,-2.5);
-      % Labels with consistent position and size
-      \node[label, text width=3cm, align=center] at (-5,-2) {Low-dimensional manifold};
-      \node[label, text width=3cm, align=center] at (0,-2) {High-dimensional noise};
-      \node[label, text width=3cm, align=center] at (5,-2) {Observed data};
-    \end{tikzpicture}
-  \end{center}
-  \caption{Manifold Learning: Parameterization vs. Embedding}\label{fig:manifolds}
-\end{figure}
-
-Within the principal manifolds framework \citep{Meng2021} -- a replicable and flexible framework for manifold learning -- the process of fitting a manifold to our data contains multiple steps.
-The key idea is that we fit a $d$-dimensional manifold to our $D$-dimensional by minimizing the sum of squares between our data and the proposed manifold.
+Within the principal manifolds framework (Meng2021) -- a replicable and flexible framework for manifold learning -- the process of fitting a manifold to our data contains multiple steps.
+The key idea of the fitting step is that we fit a $d$-dimensional manifold to our $D$-dimensional by minimizing the sum of squares between our data and the proposed manifold.
 An important extension to linear dimensionality reduction, i.e.
@@
 We note that -- although certainly intuitive -- this topological structure is not only limited to spatial abstraction, but may be extended to arbitrary dimensions of interest.
-This framework was pioneered as an extension to the PCA algorithm with curves (\citet{HastieStuetzle1989}, \citet{Tibshirani1992}) and has since found a myriad of applications in higher-dimensional extensions.
+This framework was pioneered as an extension to the PCA algorithm with curves (HastieStuetzle1989, Tibshirani1992) and has since found a myriad of applications in higher-dimensional settings.
 
-We now propose a method for incorporating prior distributional information into the principal manifolds framework.
-
-\section{Background}
-
-Consider a setting where we have fit a manifold $\mathcal{M}_d$ to our $D$-dimensional data by means of minimizing the orthogonal distance between the data and the manifold.
+Now, consider a setting where we have fit a manifold $\mathcal{M}_d$ to our $D$-dimensional data by means of minimizing the orthogonal distance between the data and the manifold.
 We consider this manifold fixed and will not touch on the fitting procedure itself.
@@
 
-where we consider each projection function $f$ to be a member of an arbitrary function space $\mathcal{F}$. We define the distance metric as the $L^2$ distance.
+where we consider each projection function $f$ to be a member of an arbitrary Sobolev space $\mathcal{F}$. We define the distance metric as the $L^2$ distance.
 
 If there exists only one projection point $f(x_i)$ for every $x_i$, every $f \in \mathcal{F}$ is one-to-one and onto ("bijective") mapping.
-We find it interesting to highlight that the projection functions in the PCA algorithm are inherently bijective, and for inferential purposes, this is a property that is often taken for granted\footnote{This is because a principal axes is a straight line, i.e.
+We find it interesting to highlight that the projection functions in the PCA algorithm are inherently bijective, and for inferential purposes, this is a property that is often taken for granted^[This is because a principal axes is a straight line, i.e.
 neither convex or concave.
-Although a point's distance to its projection may be co-minimal across $\leq 2$ dimensions, it only has one $f(x_i)$ in one principal axis.}.
+Although a point's distance to its projection may be co-minimal across $\leq 2$ dimensions, it only has one $f(x_i)$ in one principal axis.].
 In a manifold learning framework, this is no longer the case.
@@
 
-\begin{figure}
-  \begin{center}
-    \includegraphics[width=0.75\textwidth]{../fig/data-with-noise.png}
-  \end{center}
-  \caption{Noisy data projected on $\mathcal{M}_3$; the unit sphere in $\mathbb{R}^3$}\label{fig:data-with-noise}
-\end{figure}
+![Noisy data projected on $\mathcal{M}_3$; the unit sphere in $\mathbb{R}^3$](../fig/data-with-noise.png){#fig-noisy-data width=75%}
 
-\section{Directional Priors}
-
-In contrast to the radii, the $\{D-1\}$-dimensional vector of angles, $\boldsymbol{\theta_i} := [\theta_{i, 1}, \cdots, \theta_{i, D-1}]^T$ is random.
-In simple terms, $\{r_i, \boldsymbol{\theta}_i\}$ is what parameterizes the realized sample $\boldsymbol{X_i} = x_i$ in space.
-If we draw multiple samples, the random sampling variation in $\mathbf{\theta}_i$ is what captures the randomness.
-In biomedical applications, such as cancer medicine, we may have reliable prior information (i.e.
-from previous trials or expert knowledge) on directional trends of malignant growths.
-Within a Bayesian framework, using our data to update these prior directional information offers a principled, probabilistic solution to complex inference problems in settings where directionality is a key piece of information.
-
-Before we formulate our approach formally, we briefly introduce directional statistics.
-When we reparameterized our data from Cartesian coordinates into polar coordinates, we did not address the underlying probabilistic transformations.
-For example, when encoding uncertainty in a spherical setting, it may be naive to parameterize a normal density with support on $\mathbb{R}^1$ since angles are defined in the interval $[0, 2\pi]$.
-Directional statistics offer well-defined spherical reparameterizations of distributions, such as the normal.
-Instead of defining a mean vector in $\mathbb{R}^1$, we can define a mean directional vector contained in $[0, 2\pi]$.
-To account for the infinite support, the wrapped normal is defined on a support of $k\text{mod}(2\pi)$.
-Henceforth, unless stated otherwise, we will deal with the spherical parameterizations of densities to account for the spherical parameterization of our problem.
-
-\subsection{Motivating Example: 3D Sphere}
-
-We begin our discussion with an example of a 3D sphere.
-As discussed, this means that we have some simulated noisy data $\{x_i\}_{i=1}^n$ in 3D space.
-The noisy data is projected on the manifold of choice, i.e.
-the unit sphere.
-This yields the data $\{f(x_i)\}_{i=1}^n$.
-These data are parameterized in a polar coordinate system with $\{r_i, \mathbf{\theta}_i\}$, where the (non-random) radius $\vec{r_i}$ is the vector from the origin to $f(x_i)$.
-Since we are in 3D space, we have two angles.
-One angle, $\{\mathbf{\theta}\}_{i=1}^n$, which parameterizes the angle from the origin on the XY-plane.
-The angle $\{\mathbf{\phi}\}_{i=1}^n$ is the angle between the XY-plane and the point $f(x_i)$.
-
-\begin{figure}
-  \begin{center}
-    \includegraphics[width=0.75\textwidth]{../fig/projections-from-origin.png}
-  \end{center}
-  \caption{Polar parameterization with $\{r_i, \theta_i, \phi_i\}$ of projected data}\label{fig:projections-from-origin}
-\end{figure}
-
-\subsection{Formulating Directional Priors}
-
-Formulating a directional prior is not trivial.
-For one, it is important to set a reference point for both the directional prior and the angles.
-It is important that all angles are measured from the same reference point.
-For simplicity, we choose the origin.
-Now, choosing a directional prior in some direction $\mu \in [0, \pi]$ allows\footnote{Note that we choose prior means are only meaningful in $[0, \pi]$ This is because for any prior mean greater than $\pi$, we would consider the closer angle in the opposite direction.} us to incorporate information on where the projected points should be concentrated on the manifold.
-We only posit a prior on the random set $\{\theta_i\}_{i=1}^n$, expressing some prior distributional belief on the concentrations of points in the XY-plane.
-In this example, we consider the angles $\{\phi_i\}_{i=1}^n$ to be fixed, making our example univariate.
-An extension to the multivariate case would be straightforward by using the multivariate analogoues of our likelihood and prior distribution functions.
-
-Our posterior distribution on the random vector $\theta$ can be expressed as follows.
-
-\[
-\begin{aligned}
-  f_{\theta | \mu} (\theta | \mu) 
-  &\propto \left\{ \prod_{i=1}^n f_{\theta_i | \mu}(\theta_i | \mu) \right\} \cdot f_{\mu}(\mu) \\ 
-  &\propto \mathcal{L}(\theta | \mu) \cdot f_{\mu}(\mu) 
-\end{aligned}
-\]
-
-where $\boldsymbol{\mu}$ is our mean vector that we posit a prior on, i.e.
-$\mu \sim f_{\mu}(\mu)$.
-To form our likelihood, we rely on the wrapped ("spherical") normal distribution.
-The density of a wrapped normal is given by
-
-$$
-f(\theta|\mu,\sigma^2) = 
-\frac{1}{2\pi\sigma^2} \sum_{k=-\infty}^{\infty} \exp\left(-\frac{(\theta-\mu+2\pi k)^2}{2\sigma^2}\right)
-$$
-
-with the infinite support reflected in the $\mathit{mod}(k)$ argument.
-Integrating over $2\pi$ is equivalent to integrating over $\mathbb{R}$ in the canonical parameterization.
-Now, a disadvantage of using the above form is the infinite series.
-Deriving moments is not straightforward and requires (inverse) Fourier transforms.
-Thus, we will use the von Mises distribution, an analogue to the spherical normal that is easier to handle.
-The density of a von Mises distribution is given by
-
-$$
-f(\theta|\mu,\kappa) = \frac{e^{\kappa \cos(\theta-\mu)}}{2\pi I_0(\kappa)}
-$$
-
-where $\theta$ is the angle, $\mu$ is the mean direction (location parameter), $\kappa \geq 0$ is the concentration parameter, and $I_0(\kappa)$ is the modified Bessel function of the first kind of order 0.
-The von Mises distribution is also known as the circular normal distribution and is the circular analog to the normal distribution.
-As $\kappa$ increases, the distribution becomes more concentrated around the mean direction $\mu$.
-When $\kappa = 0$, the distribution reduces to the uniform distribution on the circle.
-Although we will not dive further into this, the von Mises distribution scales nicely for $p$-dimensional hyperspheres in $\mathbb{R}^p$, making the family of von Mises distributions an attractive and flexible candidate for learning tasks in spherical settings.
-
-\subsection{Variance Encodes Curvature}
-
-In the above parameterization, it is natural to ask how to define the variance in terms of the concentration parameter $\kappa$.
-Choosing the variance term naively, e.g., a constant, would of course be possible, but would negate the benefits that manifold-learning gives us as a dimensionality reduction technique which preserves local toplogical structure.
-Thus, we are interested in a principled manner of choosing the variance term that encodes the (dis-)similarity of the topological structure.
-Intuitively, if we have a directional prior on a region with little topological variation, we may be willing to update our projection more liberally.
-However, suppose that even a small change in the projection angle moves our data point into an area that is very different from our original projection.
-In that case, we not be as willing to update our projection.
-
-In manifolds that spherically parameterized, using the inverse absolute Gaussian curvature is a sensible choice to encoding variance.
-Interestingly, there is an intuitive connection to the score test in likelihood-based inference, which uses the slope at restricted MLE estimates as a measure of closeness/ similarity.
-Thus, not only does this solution make intuitive topological sense, it has a strong analogue in well-established likelihood-based methods.
-
-The Gaussian curvature of a surface can be defined as the product of the principal curvatures:
-
-$$K = k_1 \cdot k_2$$
-
-where $k_1$ and $k_2$ are the principal curvatures at a point on the surface.
-For a sphere of radius $R$, we can deduce its Gaussian curvature by considering any point on a sphere of radius $R$.
-At this point, we can find two perpendicular directions that correspond to the principal curvatures.
-Due to the perfect symmetry of a sphere, the curvature is the same in all directions at any given point.
-For a sphere, both principal curvatures are equal to $\frac{1}{R}$ (the reciprocal of the radius).
-Therefore, for our example, the Gaussian curvature is:
-
-$$K = k_1 \cdot k_2 = \frac{1}{R} \cdot \frac{1}{R} = \frac{1}{R^2}$$
-
-For a unit sphere (where $R = 1$), the Gaussian curvature\footnote{This constant Gaussian curvature of 1 is an intrinsic property of the unit sphere and is related to the fact that the total curvature integrated over the entire sphere equals $4\pi$ (by the Gauss-Bonnet theorem)} is $K = \frac{1}{1^2} = 1$.
-
-\begin{figure}
-  \begin{center}
-    \includegraphics[width=0.5\textwidth]{../fig/gaussian-curvature.png}
-  \end{center}
-  \caption{From left to right: a surface of negative Gaussian curvature (hyperboloid), a surface of zero Gaussian curvature (cylinder), and a surface of positive Gaussian curvature (sphere).}\label{fig:gaussian-curvature}
-\end{figure}
-
-\subsection{Conjugacy}
-
-A key advantage of using the von Mises distribution are conjugacy results.
-We lean on the work of \citet{mardia1976}.
-A von Mises prior distribution with mean direction $\mu$ and concentration parameter $\kappa$ (scaled by $c$) is conjugate to the likelihood.
-After observing angles $\theta_1,\ldots,\theta_n$, the posterior distribution is proportional to the following expression \citep{mardia1976}:
-
-\[
-  f(\mu_i, \mu^* | \{\theta_i\}_{i=1}^n) \propto \text{exp}(\kappa \cdot \sum_{i=1}^n \text{cos}(\theta_i - \mu_i) +\kappa^* \cdot \sum_{i=1}^n \text{cos}(\mu_i - \mu^*))  
-\]
-
-where the angles $\{\theta_i\}_{i=1}^n \sim \mathcal{VM}(\mu_i, \kappa)$ and the prior on $\mu_i$ is $\mu_i \sim \mathcal{VM}(\mu^*, \kappa^*).$\footnote{$\mathcal{VM}(\cdot, \cdot)$ denotes the von Mises distribution} In our case, we generated the data with $\mu_i = \mu = \text{circular}(0)$.
-We can also see that the above form assumes the form of shrinkage estimator\footnote{For the normal case, this procuedure leads to the James-Stein estimator \citep{guttorp1988}.}.
-The component involving $\{\theta_i\}_{i=1}^n$ is weighted by variance term $\kappa$, in our case the curvature at the projected point.
-The difference between the mean parameter of the data and the prior mean is weighted by the prior concentration parameter $\kappa^*$.
-It is interesting to extend this form of a shrinkage estimator to the explicitly topological interpretation we have mentioned above.
-If the concentration parameter of our data ($\kappa$) is low, we attribute more weight to our "prior" information.
-The above result is the desired probabilistic framework we sought to establish for this directional prior case.
-
-We are now ready to state the conjugacy result for our 3D spherical example.
-Given the following von-Mises parameterization of a prior on $\boldsymbol{\theta} \sim \mathcal{VM}(\theta_0, \tau_0 = \frac{1}{\kappa_0})$, we have the posterior mean
-
-\[
-  \frac{\tau_0}{\tau_0 + \hat{\tau}} \cdot \theta_0 + \frac{\hat{\tau}}{\tau_0 + \hat{\tau}} \cdot \bar{\theta}
-\]
-
-and with posterior variance $(\tau_0 + \hat{\tau})^{-1}$.
-Here, $\bar{\theta} := \frac{1}{n}\sum_{i=1}^n \theta_i$ and $\hat{\tau} = \frac{1}{\hat{\kappa}}$ where $\hat{\kappa}$ is the observed Gaussian curvature at a point on the manifold.
-The shrinkage toward the prior mean is immediately evident.
-
-\begin{figure}
-  \begin{center}
-    \includegraphics[width=1\textwidth]{../fig/posterior-roseplot.png}
-  \end{center}
-  \caption{Posterior distribution under prior $\theta_0$: Left $\theta_0 = \pi/2$; Right: $\theta_0 = \pi$}\label{fig:posteriors}
-\end{figure}
-
-We can see that the posterior mean is essentially the average of the prior and data means.
-This is because our manifold is a sphere with equivalent curvature at every point.
-For manifolds with heterogeneous curvature, we would see less reprojection with respect to the prior mean when curvature is more extreme.
-
-\begin{figure}
-  \begin{center}
-
-  \end{center}
-  \caption{}\label{fig:}
-\end{figure}
-
-\section{An Empirical Bayes Extension in the Service of Causality}
-
-Talk about the pre-post treatment idea. What can angles represent? Dont do any code for this, just explain.
-
-Could a prior on $\kappa$ potentially encode anticipated topological changes?
-
-\section{Limitations and Future Direction}
-
-\newpage
-
-\appendix
-\section*{Appendix A.} \label{sc:app_a}
-In this appendix we prove the following theorem from
-Section \ref{sc:intro}
-
-\newpage 
-\bibliography{references}
-
-\end{document}
+We are now ready to introduce the Bayesian framework with directional priors.

### 2025-04-10 code/03_inference.R

@@ -1,6 +1,10 @@
 # ==============================================================================
 # Title: Bayesian Inference with Spherical Normal Distribution
+# Original Author: Daniel Posmik
+# E-Mail: daniel_posmik@brown.edu
 # ==============================================================================
+
+# Set-Up =======================================================================
 library(dplyr)
 library(tidyverse)
 library(circular)

### 2025-04-10 code/04_rd.R

@@ -1,0 +1,278 @@
+# ==============================================================================
+# Title: Manifold RD Simulation
+# Original Author: Daniel Posmik
+# E-Mail: daniel_posmik@brown.edu
+# ==============================================================================
+# Set-Up =======================================================================
+library(dplyr)
+library(tidyverse)
+library(circular)
+library(plotly)
+library(mvtnorm)
+# Set working directory
+setwd("~/Documents/brown/classes/spring25/php2530-bayes/directional-priors")
+# Simulation ===================================================================
+
+# Parameters
+n <- 500
+mean_post <- pi
+mean_pre <- pi/2
+var_post <- var_pre <- 1
+kappa_post <- 0.25
+kappa_pre <- 1  
+
+# Generate samples
+set.seed(2025)
+# Pre-treatment: mean at pi/2, variance 1
+pre_treatment <- rvonmises(n, mu = circular(mean_pre), kappa = kappa_pre)
+# Post-treatment: mean at pi, variance 1
+post_treatment <- rvonmises(n, mu = circular(mean_post), kappa = kappa_post)
+
+# Calculate treatment effect distribution parameters
+mean_effect <- mean_post - mean_pre
+# Variance formula as specified
+variance_effect <- sqrt((kappa_pre * cos(mean_pre) + kappa_post * cos(mean_post))^2 + 
+                          (kappa_pre * sin(mean_pre) + kappa_post * sin(mean_post))^2)
+
+# Calculate treatment effect samples
+# For circular data, we compute the difference between post and pre angles
+# We need to wrap the angles to ensure they stay in [0, 2π]
+treatment_effect <- (as.numeric(post_treatment) - as.numeric(pre_treatment)) %% (2*pi)
+
+# Function to create rose plot with plotly
+create_rose_plot_style <- function(angles, mean_angle, title, kappa_or_var = 1, is_variance = FALSE, 
+                                   plot_color = 'rgba(200, 200, 255, 0.7)', line_color = 'darkblue') {
+  # Number of bins for the rose plot
+  n_bins <- 24
+  bin_width <- 2*pi/n_bins
+  
+  # Create bins for data
+  breaks <- seq(0, 2*pi, length.out = n_bins + 1)
+  bin_mids <- breaks[-1] - bin_width/2
+  
[…]

### 2025-04-30 output/paper.tex
Added README
First draft done, needs a beauty overhaul though

@@
 \usepackage{subcaption}
+\usepackage{hyperref}
 \definecolor{darkorange}{RGB}{204,85,0}
@@
 %% short title here
-Directional Priors in Manifolds
+Toward a Manifold RD Estimator
 }{
@@
 
-\title{	Directional Priors in Manifold Learning: A Causal Perspective \\
+\title{Bayesian Topological Weights in Causal Inference -- \\ Toward a Manifold RD Estimator \\
 \vspace{.1in}
@@
 
+\textcolor{red}{There's a parellel to random effects; Call them Bayesian topo weights; Algorithm: while $i < N$, 1) Draw $\theta_i$ from posterior, 2) get $kappa_i$ at that $\theta_i$, 3) Calculate weight}
+
 Manifold learning is dimensionality reduction technique that has proven useful in settings where data are high-dimensional and non-linear.
@@
   \begin{center}
-    \begin{tikzpicture}[scale=0.85]
+    \begin{tikzpicture}[scale=0.75]
       % Define styles
@@
 For a unit sphere, the Gaussian curvature is $K = 1$, an intrinsic property related to the total curvature integrated over the entire sphere equaling $4\pi$ by the Gauss-Bonnet theorem.
-
-\begin{figure}[h!]
-  \begin{center}
-    \includegraphics[width=0.5\textwidth]{../fig/gaussian-curvature.png}
-  \end{center}
-  \caption{From left to right: a surface of negative Gaussian curvature (hyperboloid), a surface of zero Gaussian curvature (cylinder), and a surface of positive Gaussian curvature (sphere).}\label{fig:gaussian-curvature}
-\end{figure}
-
 This curvature-based approach provides a meaningful way to weight observations based on their location in the manifold's topological structure.
@@
 
-\textcolor{red}{Make it crystal clear that the posterior variance is the Bayesian thing here!!}
+Our goal is to propose an estimator that exploits topological information from pre- and post-treatment manifolds.
+Before we begin, recall the definition of a regression discontinuity (RD) estimator when treatment is administered at time $t^*$.
 
-An interesting application of directional priors is a causal setting where we seek to estimate treatment effects after some treatment has been administered.
-A popular class of estimators for comparing changes in outcomes of interest close to a treatment cut-off are regression discontinuity (RD) estimators.
-RD estimators exploit the the balance between units close to the treatment cut-off, assuming that treatment does not depend on any unit characteristics.
-Recovering causal effects in RD estimators hinges on the assumption that in terms of unit characteristics, treatment assignment is quasi-random very close to the cut-off.
-Popular examples in literature include measuring the effect of advanced coursework on students above some cutoff, comparing their outcomes to students just below said cutoff.
+\[
+  \tau(t^*) = \lim_{\delta \rightarrow 0} \mathbb{E}[\theta | Z = 1, t = t^* + \delta] - \lim_{\delta \rightarrow 0} \mathbb{E}[\theta | Z = 0, t = t^* - \delta] 
+\]
 
-In the principal manifold learning setup, directional priors offer a promising Bayesian approach to causal inference.
-For simplicity, assume that treatment is administered at time point $t^*$.
-Then, all units in $(-\infty, t^*)$ are pre-treatment units, and all units in $[t^*, \infty)$ are post-treatment units.
-In a canonical RD setting, under a set of identification assumptions, we can recover an average treatment by considering values of $Y_i$ immediately to the right (say $Y_{it} |_{t^* + \delta}$) and left ($Y_{it} |_{t^* - \delta}$) of the cutoff.
-This yields the causal estimate
+As it relates to the principal manifolds framework, after treatment happens at $t=t^*$, we can fit the post-treatment manifold $\mathcal{M}_d^{t \geq t^*}$.
+In the spirit of RD-based causal estimation, we want to emphasize that $\mathcal{M}_d^{t \geq t^*}$ is \textit{not} fit iteratively for each post-treatment unit, i.e.
+we do not obtain the post-treatment unit by combining the post-treatment units with pre-treatment units.
+Instead, we fit two manifolds with disjoint pre- and post-treatment data.\footnote{That being said, it may be interesting to let the choice of manifold (e.g., the shape, highest order, etc.) from the pre-treatment case inform the post-treatment case, effectively borrowing prior topological information on a more global level.
+This is a really interesting direction in the Bayesian paradigm and touches on the notion of persistent homology.
+Although this discussion goes beyond the scope of this paper, it highlights how nuanced various levels/ sources of topological information can be.} Namely, we observe the outcomes of $m$ additional units in the immediate post-treatment period, and fit $\mathcal{M}_d^{t \geq t^*}$ with data $\{\theta_{t}\}_{i=n+1}^{m}|_{t \geq t^*}$.
+The mean projection angle of $\mathcal{M}_d^{t \geq t^*}$, $\bar{\theta}^{t \geq t^*}$, is the post-treatment mean projection angle.
+It reflects the directionality under treatment.
+The following figure presents this admittedly complex idea visually.
 
-$$
-\hat{\tau}= \bar{Y}|_{t = t^* + \delta} - \bar{Y}|_{t = t^* - \delta} 
-$$
-
-The above estimator is commonly measured in the Local Linear Regression (LLR) framework, i.e.
-measure the ATE as the difference in post- and pre-treatment means, given that they are sufficiently close to the treatment cutoff $t^*$.
-\citet{Branson2019} considers a Bayesian RD setting where a Gaussian process prior is placed on the pre- and post-treatment mean response functions.
-We believe this work, although not directly related to the use of topological information in causal inference, encapsulates the desire for more flexible, data-driven, and principled causal inference through Bayesian paradigms.
-
-\subsection{Intuition}
-
-In a principal manifolds framework, we can also exploit this notion of quasi-randomness to recover causal effects.
-Specifically, we propose the estimation of causal treatment effects using the posterior density of projection angles with empirical prior variances encoding the local topological structure of pre-treatment manifolds.
-For this procedure, suppose we use the set of pre-treatment units $\{Y_{it}\}_{i=1}^n|_{t < t^*}$ to fit the $d$-dimensional manifold $\mathcal{M}_d^{t < t^*}$.
-As we have shown in Section~\ref{sc:directional-priors}, a polar reparameterization allows us to recover the projection angles $\{\theta_i\}$ as well as their mean which we shall denote as $\bar{\theta}^{t < t^*}$.
-This mean pre-treatment projection angle $\bar{\theta}^{t < t^*}$ has a curvature which we will denote as $\kappa^{t<t^*}$.
-Together, the mean pre-treatment projection angle and curvature sufficiently parameterize the first two moments of the pre-treatment outcome of interest.
-
-After treatment happens at $t=t^*$, we can fit the post-treatment manifold $\mathcal{M}_d^{t \geq t^*}$.
-In the spirit of LLR-based causal estimation, we want to emphasize that $\mathcal{M}_d^{t \geq t^*}$ is *not* fit iteratively for each post-treatment unit, i.e.
-we do not obtain the post-treatment unit by combining the post-treatment units with pre-treatment units.
-Instead, our pre- and post-treatment manifolds are topological analogues to $\bar{Y}|_{t = t^* - \delta}$ and $\bar{Y}|_{t = t^* + \delta}$, respectively.
-That being said, it may be interesting to let the choice of manifold (e.g., the shape, highest order, etc.) from the pre-treatment case inform the post-treatment case, effectively borrowing prior topological information on a more global level.
-\footnote{This is a really interesting direction in the Bayesian paradigm and touches on the notion of persistent homology.
-Although this discussion goes beyond the scope of this paper, it highlights how nuanced various levels/ sources of topological information can be.} Now, suppose that we observe the outcomes of $m$ additional units in the immediate post-treatment period, then the data $\{Y_{it}\}_{i=n+1}^{m}|_{t \geq t^*}$ allows us to fit the manifold $\mathcal{M}_d^{t \geq t^*}$.
-The mean projection angle of $\mathcal{M}_d^{t \geq t^*}$, i.e.
-$\bar{\theta}^{t \geq t^*}$ will update from $\bar{\theta}^{t < t^*}$, and yield a new posterior estimate for the projection angle.
-
-\begin{figure}
+\begin{figure}[h!]
   \begin{center}
@@
 
-\subsection{The Estimator}
+\subsection{The Manifold RD Estimator}
 
-Combining our insights from RD estimators with the conjugacy results of Section~\ref{sc:conjugacy}, we propose a Bayesian manifold RD estimator in its simplest form.
-After having the pre- and post-treatment manifold, consider subtracting the pre-treatment distribution of angles from the post-treatment distribution.
-This yields a RD estimator of the form:
+It is common to estimate the causal effect, write $\hat{\tau}(t^*)$, using a weighting approach \textcolor{red}{cite these bad bois}.
+Units that are located closer to $t^*$ are weighted more, while units that are located further away are weighted less.
+This makes sense in the context of the identification assumption underlying the RD estimator, i.e.
+units close to the cutoff only differ in treatment assignment.
+Thus, treatment is considered quasi-random.
+
+Our crux is to translate this weighting scheme into a topological setting.
+Intuitively, we want to achieve the following: For each additional unit we observe after the treatment has been administered, evaluate whether this observation is \textit{topologically similar} to our pre-treatment mean.
+If it is, we want to assign a larger weight to it.
+If it is not, we want to assign less weight to it.
+Remember that we formalized the notion of topological similarity using the curvature $\kappa|\theta_i$, i.e.
+for unit $i$, the curvature at the projection angle encodes its topological neighborhood.
+Thus, if some unit is embedded close to what we would expect from our pre-treatment data, we want to assign a lot of weight to it since it is topologically similar.
+
+This thought process yields the following weighting approach. Let $\hat{\tau}(t^*)$ be the weighted treatment effect estimator,
+
+\[
+  \hat{\tau}(t^*) = \sum_{j=n+1}^{m} \frac{w_j \cdot \theta_j^{t \geq t^*}}{m - n} - \sum_{i=1}^{n} \frac{w_i \cdot \theta_i^{t < t^*}}{n}
+\]
+
+For all pretreatment observations, $i \in \{1, \cdots, n\}$, the weight $w_i$ simply quantifies the difference of unit $i$'s projection angle $\theta_i$ from the mean pre-treatment projection angle $\bar{\theta}^{t < t^*}$.
+This is analogous to the canonical weighting scheme in literature.
+When we enter the post-treatment period, with observations $j \in \{n+1, \cdots, m\}$, the weight $w_j$ is drawn from a posterior with prior around the pre-treatment projection angle.
+Its likelihood is updated for each additional point in the post-treatment period.
+Thus, we formally define our weight function as follows:
 
 $$
-f_{\tau | \theta} = 
-\mathcal{VM}
-\left(
-  \bar{\theta}^{t \geq t^*} - \bar{\theta}^{t < t^*}, \sigma^2 
-\right)
+w_i(\kappa|\theta) = 
+  \frac{1}{\|\kappa_i^{t<t^*}|\textcolor{darkorange}{\{\theta_i\}_{i=1}^{t}} - \kappa^{t<t^*}|\bar{\theta}^{t<t^*}\|}
 $$
+
+Here, $\kappa^{t<t^*}|\theta_i$ denotes the Gaussian curvature on the pre-treatment manifold, evaluated at $\theta_i$.
+Meanwhile, $\kappa^{t<t^*}|\bar{\theta}^{t<t^*}$ denotes the Gaussian curvature on the pre-treatment manifold, evaluated at the mean pre-treatment projection angle.
+In the post-treatment period, this quantity is always non-random.
+
+Before we proceed, we want to emphasize the difference in interpretation of ${\{\theta_i\}_{i=1}^t}$ from the pre-treatment to the post-treatment case.
+In the pre-treatment case, ${\{\theta_i\}_{i=1}^t}$ represents the prior distribution using pre-treatment observations.
+In the post-treatment case, ${\{\theta_i\}_{i=1}^t}$ becomes a posterior distribution, where the prior distribution is updated with each post-treatment observation in $j \in \{n+1, \cdots, t\}$.
+This implies that we can express our data vector as a disjoint union of sets:
+
+$$
+\textcolor{darkorange}{\{\theta_i\}_{i=1}^t} = 
+\begin{cases}
+\{\theta_i\}_{i=1}^{t}, & \text{if } t < t^* \\ 
+\{\theta_i\}_{i=1}^{n} \cup \{\theta_j\}_{j=n+1}^{t}, & \text{if } t \geq t^*
+\end{cases}
+$$
+
+In the pre-treatment case, we have a canonical von Mises distribution.
+In the post-treatment case, we make use of our conjugacy result from Section~\ref{sc:directional-priors}.
+Naturally, in a causal setting, we will focus on the case $t \geq t^*$, i.e.
+when post-treatment data is available.
+This yields our hierarchical manifold RD estimator:
+
+\[
+  \begin{aligned}
+    \tau | &w_i \\
+           &w_i = w_i(\kappa^{t<t^*}|\theta_i) \\
+           &\theta_i \sim \mathcal{VM}(\mu, \sigma^2)
+\end{aligned}
+\]
 
@@
 \[
-  \sigma^2 = \sqrt{\left[\kappa^{t < t^*}\cos(\bar{\theta}^{t < t^*}) + \textcolor{darkorange}{\kappa^{t \geq t^*}\cos(\bar{\theta}^{t \geq t^*})}\right]^2 
+\begin{aligned}
+  &\mu =  \frac{1/\kappa^{t<t^*}}{1/\kappa^{t<t^*} + 1/\kappa^{t \geq t^*}} \cdot \bar{\theta}^{t < t^*} + \frac{1/\kappa^{t \geq t^*}}{1/\kappa^{t < t^*} + 1/\kappa^{t \geq t^*}} \cdot \bar{\theta}^{t \geq t^*} \\
+  &\sigma^2 = \sqrt{\left[\kappa^{t < t^*}\cos(\bar{\theta}^{t < t^*}) + \textcolor{darkorange}{\kappa^{t \geq t^*}\cos(\bar{\theta}^{t \geq t^*})}\right]^2 
   + 
 \left[\kappa^{t < t^*}\sin(\bar{\theta}^{t < t^*}) + \textcolor{darkorange}{\kappa^{t \geq t^*}\sin(\bar{\theta}^{t \geq t^*})}\right]^2} 
+\end{aligned}
 \]
@@
 
-Here, we estimate the Average Treatment Effect (ATE) as the difference between the mean projection angles of the post- and pre-treatment manifold, respectively.
-As illustrated in Figure~\ref{fig:manifold-rd}, this difference captures the directional shift in data geometry following treatment intervention.
-The posterior variance, expressed as the Euclidean norm of the difference between weighted directional vectors, quantifies our uncertainty about this effect.
-Intuitively, a larger norm indicates greater topological dissimilarity between pre- and post-treatment manifolds, suggesting a more substantial causal impact.
+We want to emphasize that the Bayesian inference procedure is not the causal estimator $\hat{\tau}$ directly.
+Rather the weights as a function of $\kappa$ (evaluated at $\theta$).
+Posterior inference through the above mechanism is possible as soon as we observe at least one post-treatment unit.
+An immediate advantage of this estimation procedure is that we can update our estimates readily for each new observation.
 
-To illustrate this, let us recycle the toy example from Section~\ref{sc:conjugacy}.
+The manifold RD estimator is similar to weighted RD estimators in literature.
+A common and intuitive estimation procedure for RD estimators is the Local Linear Regression (LLR) framework, i.e.
+measure the LATE as the difference in post- and pre-treatment means, given that they are sufficiently close to the treatment cutoff $t^*$.
+\citet{Branson2019} considers a Bayesian RD setting where a Gaussian process prior is placed on the pre- and post-treatment mean response functions.
+We believe this work, although not directly related to the use of topological information in causal inference, encapsulates the desire for more flexible, data-driven, and principled causal inference through Bayesian paradigms.
+
+\subsection{Estimator Properties and Simulation}
+
+An analytical derivation of our estimator's properties is simplified greatly by our conjugacy results.
+Naturally, outside of the von Mises parameterization, estimator properties can be determined by posterior sampling and numerical methods.
+To provide further insight into the estimator, we will analytically derive its first two moments and provide a brief simulation.
+
+The expectation of $\hat{\tau}(t^*)$, where we let $w_{i,j} = w_{i,j}(\kappa|\mu)$ denote the respective weights, is the following:
+\[
+  \begin{aligned}
+    \mathbb{E}(\hat{\tau}(t^*)) &= \sum_{j=n+1}^{m} \frac{w_j \cdot \theta_j^{t \geq t^*}}{w_j} - \sum_{i=1}^{n} \frac{w_i \cdot \theta_i^{t < t^*}}{w_i}
+  \end{aligned}
+\]
+
+Intuitively, we see that our manifold RD estimator is a topologically weighted average of post-treatment observations, that accounts for the pre-treatment "baseline". For our variance,
+
+$$
+\begin{aligned}
+\mathbb{V}(\hat{\tau}(t^*)) &= \mathbb{V}\left(\sum_{j=n+1}^{m} \frac{w_j \cdot \theta_j^{t \geq t^*}}{w_j} - \sum_{i=1}^{n} \frac{w_i \cdot \theta_i^{t < t^*}}{w_i}\right) \\
+                            &= \mathbb{V}\left(\sum_{j=n+1}^{m} \frac{w_j \cdot \theta_j^{t \geq t^*}}{w_j}\right) + \mathbb{V}\left(\sum_{i=1}^{n} \frac{w_i \cdot \theta_i^{t < t^*}}{w_i}\right) \\
+                            &\underbrace{- 2\operatorname{Cov}\left(\sum_{j=n+1}^{m} \frac{w_j \cdot \theta_j^{t \geq t^*}}{w_j}, \sum_{i=1}^{n} \frac{w_i \cdot \theta_i^{t < t^*}}{w_i}\right)}_{=0}
+\end{aligned}
+$$
+
+Since the pre-treatment and post-treatment observations are assumed to be independent, the covariance term equals zero.
+The evaluation of the variance term depends on the underlying curvature of the manifold, and is thus highly dependent on the choice of manifold.
+
+Returning to our spherical example, our moment expressions simplify greatly.
+This is because in a sphere, our curvature is constant everywhere, thus yielding no variation in the weights.
+Therefore, our manifold RD estimator reduces to our canonical RD estimator, i.e.
+the difference of post- and pre-treatment angles.
+
+$$
+\begin{aligned}
+  \mathbb{E}(\hat{\tau}(t^*)) &= \mu^{t \geq t^*} - \mu^{t < t^*} \\ 
+  \mathbb{V}(\hat{\tau}(t^*)) &= \frac{(\kappa^{t \geq t^*})^{-1}}{n-m} + \frac{(\kappa^{t < t^*})^{-1}}{n} 
+\end{aligned}
+$$
+
+Although the spherical example is thus somewhat trivial, we want to present a simulation study of the manifold RD estimator when both manifolds are spheres.
+We recycle the toy example from Section~\ref{sc:conjugacy}.
 Suppose we have a post-treatment mean projection angle, $\bar{\theta}^{t \geq t^*} = \pi$ and a pre-treatment mean projection angle, $\bar{\theta}^{t < t^*} = \pi/2$.
@@
 Thus, the manifold we have fit grows from a unit circle (radius = $1$) to a post-treatment circle with radius $2$.
+This is relevant to biomedical settings where we believe our object of interest, e.g., a tumor growth, grows equally in size after treatment.
 This means that while the pre-treatment Gaussian curvature stays 1, i.e.
@@
   \begin{center}
-    \includegraphics[width=1\textwidth]{../fig/rd-sim.png}
+    \includegraphics[width=0.95\textwidth]{../fig/rd_sim.png}
   \end{center}
-  \caption{Simulation of Treatment Effect $\hat{\tau}$}\label{fig:rd-sim}
+  \caption{Distribution of Pre- and Post-Treatment Data, Treatment Effect, and Estimated Treatment Effect}\label{fig:rd-sim}
 \end{figure}
 
-We can see that the resulting treatment effect estimate $\hat{\tau}$ is indeed very close to $\frac{\pi}{2}$. The variance term is slightly above 1, i.e. greater than the individual parts.
+We can see that our estimator captures the treatment effect well.
+We want to reiterate that in the spherical setting, the manifold RD estimator simplifies to the canonical estimator.
+That being said, it is reassuring to see well-behaved simulation-based results in this simplest of settings.
+
+\section{Discussion and Future Direction}
 
@@
 This formulation naturally preserves identification assumptions required for causal interpretation while accommodating the topological abstractions within the principal manifold framework.
-The magnitude of the posterior variance directly reflects how drastically treatment has altered the local geometry of our outcome space, e.g., if we have a large difference, we want to encode that topological uncertainty into our ATE.
+The magnitude of the posterior variance directly reflects how drastically treatment has altered the local geometry of our outcome space, e.g., if we have a large difference, we want to encode that topological uncertainty into our LATE.
 
@@
 Namely, we are not simply updating our fitted manifold with the new data, and letting the posterior angle update iteratively with each new post-treatment observation.
+This is certainly an interesting angle in the context of statistical learning procedures (e.g., prediction tasks), but would complicate causal identification.
 Rather, we are using the pre-treatment manifold to borrow topological information for our post-treatment mean vector.
@@
 We round out this section with an appealing information-theoretic point of view.
-This manifold RD estimate is fundamentally analogous to the existing methods.
+This manifold RD estimate is fundamentally analogous to the existing methods that penalize observations for being further away from treatment onset in time.
 The main difference is that this procedure accounts for the topological information of our embedding space.
 Existing methods are quick to discard this information and assume it away.
-Principled manifold learning approaches combined with the flexible probabilistic advantages of Bayesian methods can therefore yield estimates that respect both prior and topological information, resulting in more flexible and data-driven inference.
-
-\section{Limitations and Future Direction}
+Through our Bayesian topological weights, we now have a procedure for iteratively updating our beliefs in the underlying topological structure of our data.
+Each additional observation in the post-treatment period allows us to strengthen our beliefs in local topological changes.
 
 Future work could address the use of priors for the construction of the mean projection vectors in the pre- and post-treatment periods, respectively.
+That is, within the existing manifold RD framework, posit hyperpriors within each treatment period.
+This connects neatly to the idea of encoding global manifold prior information.
+A promising avenue within the broader topological data analysis is mmanifold learning drawing from persistent homology.
+To the author's knowledge, there currently exists no method to obtain causal estimates in a manifold RD setup.
+With that in mind, this idea is rudimentary and can benefit from more rigorous analysis and robust theoretical guarantees.
+
+\section{Code and Data Availability Statement}
+
+Code necessary to reproduce the simulation study in this paper is available at the GitHub repository: \href{https://github.com/posmikdc/directional-priors}{https://github.com/posmikdc/directional-priors}
 
@@
 
-\appendix
-\section*{Appendix A.} \label{sc:app_a}
-All code and documentation can be found here. [Insert Link]
-
-\newpage 
+%\appendix
+%\section*{Appendix A.} \label{sc:app_a}
+%\newpage 
 \bibliography{references}

### 2025-04-30 code/03_inference.R

@@ -248,3 +248,5 @@ combined_plot <- plotly::subplot(plot1, plot2, nrows = 1, shareY = TRUE) %>%
 
 # Display the combined plot
 combined_plot
+
+ggsave("fig/posterior-roseplot.png", combined_plot, width = 12, height = 6, dpi = 300)

### 2025-04-30 code/04_rd.R

@@ -1,278 +1,424 @@
 # ==============================================================================
-# Title: Manifold RD Simulation
+# Title: Manifold RD Estimator Simulations
 # Original Author: Daniel Posmik
 # E-Mail: daniel_posmik@brown.edu
 # ==============================================================================
-# Set-Up =======================================================================
+
+# Setup ========================================================================
 library(dplyr)
-library(tidyverse)
+library(ggplot2)
 library(circular)
-library(plotly)
-library(mvtnorm)
+library(latex2exp)
+
 # Set working directory
 setwd("~/Documents/brown/classes/spring25/php2530-bayes/directional-priors")
-# Simulation ===================================================================
 
-# Parameters
-n <- 500
-mean_post <- pi
+# Set Seed
+set.seed(1)
+
+# Simulation Parameters ========================================================
+n_sims <- 500       # Number of simulation runs
+n_pre <- 500        # Number of pre-treatment observations
+n_post <- 500       # Number of post-treatment observations
+t_star <- n_pre     # Cutoff point (time of treatment)
+
+# Pre-treatment parameters (unit circle)
 mean_pre <- pi/2
-var_post <- var_pre <- 1
-kappa_post <- 0.25
-kappa_pre <- 1  
+kappa_pre <- 5      # Concentration parameter (inverse of variance)
+radius_pre <- 1     # Radius of pre-treatment manifold (unit circle)
 
-# Generate samples
-set.seed(2025)
-# Pre-treatment: mean at pi/2, variance 1
-pre_treatment <- rvonmises(n, mu = circular(mean_pre), kappa = kappa_pre)
-# Post-treatment: mean at pi, variance 1
-post_treatment <- rvonmises(n, mu = circular(mean_post), kappa = kappa_post)
+# Post-treatment parameters (larger circle)
+mean_post <- pi
+kappa_post <- 5     # Same concentration for fair comparison
+radius_post <- 2    # Radius of post-treatment manifold
 
-# Calculate treatment effect distribution parameters
-mean_effect <- mean_post - mean_pre
-# Variance formula as specified
-variance_effect <- sqrt((kappa_pre * cos(mean_pre) + kappa_post * cos(mean_post))^2 + 
[…]

### 2025-05-09 output/paper.tex
Everything finalized, ready for submission

@@
 
-\title{Bayesian Topological Weights in Causal Inference -- \\ Toward a Manifold RD Estimator \\
+\title{Bayesian Topological Weights in Causal Inference -- \\ Toward a Manifold Regression Discontinuity Estimator \\
 \vspace{.1in}
@@
 
-\textcolor{red}{There's a parellel to random effects; Call them Bayesian topo weights; Algorithm: while $i < N$, 1) Draw $\theta_i$ from posterior, 2) get $kappa_i$ at that $\theta_i$, 3) Calculate weight}
-
 Manifold learning is dimensionality reduction technique that has proven useful in settings where data are high-dimensional and non-linear.
-Often, manifold learning algorithms are used when the topological structure of the data are to be preserved in a statistical learning task\footnote{For an introduction, see \citet{Meila2023}}.
+Often, manifold learning algorithms are used when the topological structure of the data are to be preserved in a statistical learning task\footnote{For a suitable introduction, see \citet{Meila2023}}.
 Within the manifold learning framework, data are assumed to live on a lower dimensional manifold and are corrupted by high-dimensional noise.
@@
 
-Within the principal manifolds framework \citep{Meng2021} -- a replicable and flexible framework for manifold learning -- the process of fitting a manifold to our data contains multiple steps.
-The key idea of the fitting step is that we fit a $d$-dimensional manifold to our $D$-dimensional by minimizing the sum of squares between our data and the proposed manifold.
-An important extension to linear dimensionality reduction, i.e.
-the principal components algorithm (PCA), is that we allow our proposed manifold to preserve underlying topological structure of our data.
+Within the principal manifolds framework \citep{Meng2021} -- a replicable and flexible framework for manifold learning -- we effectively extend linear dimensionality reduction, i.e.
+the principal components algorithm (PCA), by allowing our proposed manifold to preserve underlying topological structure of our data.
 In a way, manifold learning reduces the dimensionality of data with an explicit focus on the topology of it.
@@
 
-We now propose a method for incorporating prior distributional information into the principal manifolds framework.
+We propose a causal regression discontinuity (RD) method within the manifold framework.
+To effectively account for the change in underlying topological structure, we propose Bayesian weights that incorporate prior distributional information from the pre-treatment period.
+Before we dive into the causal estimator, we provide important background on the concept of directional priors in manifolds in the next section.
+The paper is rounded by a brief presentation, simulation, and discussion on the manifold RD estimator.
 
-\section{Background}\label{sc:background}
+\section{Manifold Learning and Directional Information}\label{sc:background}
 
@@
 
-where we consider each projection function $f$ to be a member of an arbitrary Sobolev space $\mathcal{F}$. We define the distance metric as the $L^2$ distance.
+where we consider each projection function $f$ to be a member of an arbitrary function space $\mathcal{F}$ (for convenience, assume $\mathcal{F}$ is a Sobolev space).
+We define the distance metric as the $L^2$ distance.
 
@@
 
-Now, given the data $[\{x_i\}_{i=1}^n, \{f(x_i)\}_{i=1}^n]$, we can reparameterize our space into polar coordinates to obtain a vector representation of the collection $f \in \mathcal{F}$.
+\subsection{Directional Priors}\label{sc:directional-priors}
+
+Given the projected data $[\{x_i\}_{i=1}^n, \{f(x_i)\}_{i=1}^n]$ on the manifold, we can reparameterize our space into polar coordinates to obtain a vector representation of the collection $f \in \mathcal{F}$.
 Converting a Cartesian parameterization in space with $D$ dimensions into polar coordinates yields the $d$-dimensional vector $[r_i^*; \theta_{i, 1}, \cdots, \theta_{i, D-1}]$, i.e.
@@
 
-\section{Directional Priors}\label{sc:directional-priors}
-
-In contrast to the radii, the $\{D-1\}$-dimensional vector of angles, $\boldsymbol{\theta_i} := [\theta_{i, 1}, \cdots, \theta_{i, D-1}]^T$ is random.
-In simple terms, $\{r_i, \boldsymbol{\theta}_i\}$ is what parameterizes the realized sample $\boldsymbol{X_i} = x_i$ in space.
+In contrast to the radii, the $\{D-1\}$-dimensional vector of angles, $\theta_i := [\theta_{i, 1}, \cdots, \theta_{i, D-1}]^T$ is random.
+In simple terms, $\{r_i, \theta_i\}$ is what parameterizes the realized sample $X_i = x_i$ in space.
 If we draw multiple samples, the random sampling variation in $\mathbf{\theta}_i$ is what captures the randomness.
@@
 
-Before we formulate our approach formally, we briefly introduce directional statistics.
-When we reparameterized our data from Cartesian coordinates into polar coordinates, we did not address the underlying probabilistic transformations.
-For example, when encoding uncertainty in a spherical setting, it may be naive to parameterize a normal density with support on $\mathbb{R}^1$ since angles are defined in the interval $[0, 2\pi]$.
-Directional statistics offer well-defined spherical reparameterizations of distributions, such as the normal.
-Instead of defining a mean vector in $\mathbb{R}^1$, we can define a mean directional vector contained in $[0, 2\pi]$.
-To account for the infinite support, the wrapped normal is defined on a support of $k\text{mod}(2\pi)$.
-Henceforth, unless stated otherwise, we will deal with the spherical parameterizations of densities to account for the spherical parameterization of our problem.
-
 \subsection{Motivating Example: 3D Sphere}
 
-We begin our discussion with an example of a 3D sphere.
-As discussed, this means that we have some simulated noisy data $\{x_i\}_{i=1}^n$ in 3D space.
-The noisy data is projected on the manifold of choice, i.e.
-the unit sphere.
+For the remainder of this paper, we will stick to the example of a 3D sphere.
+The noisy data is projected on the sphere of choice, for simplicity we choose the unit sphere.
 This yields the data $\{f(x_i)\}_{i=1}^n$.
@@
 
-Formulating a directional prior requires careful consideration of reference points.
-All angles must be measured from the same reference point—for simplicity, we choose the origin.
+We are now ready to formally define directional priors.
 A directional prior with mean direction $\mu \in [0, \pi]$ allows us to incorporate information about where projected points should concentrate on the manifold.
@@
 
-Our posterior distribution on the random vector $\theta$ can be expressed as:
+Our posterior distribution on the random vector $\theta$ can be expressed as follows:
 
@@
   f_{\theta | \mu} (\theta | \mu) 
-  &\propto \left\{ \prod_{i=1}^n f_{\theta_i | \mu}(\theta_i | \mu) \right\} \cdot f_{\mu}(\mu) \\ 
-  &\propto \mathcal{L}(\theta | \mu) \cdot f_{\mu}(\mu) 
+  &\propto \left\{ \prod_{i=1}^n f_{\theta_i | \mu}(\theta_i | \mu) \right\} \cdot f_{\mu}(\mu)
 \end{aligned}
@@
 
-For our concentration parameter $\kappa$, naive selection (e.g., a constant) would be possible but would negate the benefits of manifold learning as a topology-preserving dimensionality reduction technique.
-Instead, we propose a principled approach that encodes local topological structure into the variance parameter.
+In smooth manifolds, choosing the inverse Gaussian curvature as our variance term yields a principled approach to encode topological structure into the variance paramater.
 Intuitively, in regions with little topological variation, we might update our projection more liberally.
 Conversely, if a small change in projection angle moves our data point into a topologically distinct region, we would be more conservative with updates.
-
-In spherically parameterized manifolds, the inverse absolute Gaussian curvature provides a natural variance encoding.
 This approach has an elegant connection to the score test in likelihood-based inference, which uses the slope at restricted MLE estimates as a similarity measure.
-Thus, our solution has both topological intuition and established statistical parallels in traditional inference methods.
+Considering $\kappa$ as random nuisance parameter also provides a bridge to the mixed models literature.
 
@@
 
-For our 3D spherical example, given a prior $\boldsymbol{\theta} \sim \mathcal{VM}(\theta_0, \tau_0 = \frac{1}{\kappa_0})$, the posterior mean is:
-
-$$
-\underbrace{\frac{\tau_0}{\tau_0 + \hat{\tau}} \cdot \theta_0}_{\text{Weighted Prior Vector}} + \underbrace{\frac{\hat{\tau}}{\tau_0 + \hat{\tau}} \cdot \bar{\theta}}_{\text{Weighted Data Vector}}
-$$
-
-with posterior variance $(\tau_0 + \hat{\tau})^{-1}$, where $\bar{\theta} := \frac{1}{n}\sum_{i=1}^n \theta_i$ and $\hat{\tau} = \frac{1}{\hat{\kappa}}$ with $\hat{\kappa}$ being the observed Gaussian curvature at the manifold point.
-The shrinkage toward the prior mean is immediately evident in this expression.
-
 \begin{figure}[h!]
   \begin{center}
-    \includegraphics[width=1\textwidth]{../fig/posterior-roseplot.png}
+    \includegraphics[width=0.85\textwidth]{../fig/posterior-roseplot.png}
   \end{center}
@@
 
-It is common to estimate the causal effect, write $\hat{\tau}(t^*)$, using a weighting approach \textcolor{red}{cite these bad bois}.
+It is common to estimate the causal effect, write $\hat{\tau}(t^*)$, using a weighting approach.
 Units that are located closer to $t^*$ are weighted more, while units that are located further away are weighted less.
